## Activity 3:  IMDb Movie Reviews --- Displaying Quantitative Variables

\setstretch{1}

### Learning outcomes

* Identify and create appropriate summary statistics and plots
  given a data set or research question for quantitative data.

* Interpret the following summary statistics in context:
  median, lower quartile, upper quartile, 
  standard deviation, interquartile range.

### Terminology review

In today's activity, we will review summary measures and plots for quantitative variables.  Some terms covered in this activity are:

* Two measures of center: mean, median

* Two measures of spread (variability): standard deviation, interquartile range (IQR)

* Types of graphs: box plots, dot plots, histograms

* Identify and create appropriate summary statistics and plots given a data set or research question for a single categorical and a single quantitative variable.

* Interpret the following summary statistics in context:
  median, lower quartile, upper quartile,
  standard deviation, interquartile range.

* Given a plot or set of plots, describe and compare the distribution(s)
  of a single quantitative variable
  (center, variability, shape, outliers).
  

To review these concepts, see Chapter 5 in the textbook.

### Movies released in 2016

A data set was collected on movies released in 2016 [@imdb].  Here is a list of some of the variables collected on the observational units, movies released in 2016.  

| **Variable** 	| **Description** |
|----	|-------------	|
| `budget_mil` | Amount of money (in US $ millions) budgeted for the production of the movie |
| `revenue_mil` | Amount of money (in US $ millions) the movie made after release|
| `duration` | Length of the movie (in minutes)|
| `content_rating` | Rating of the movie (`G`, `PG`, `PG-13`, `R`, `Not Rated`)|
| `imdb_score` | IMDb user rating score from 1 to 10 |
| `genres` | Categories the movie falls into (e.g., Action, Drama, etc.) |
| `facebook_likes` | Number of likes a movie receives on Facebook |

\newpage

#### Summarizing a single quantitative variable {-}

The `favstats()` function from the `mosaic` package gives the summary statistics for a quantitative variable. The `R` output below provides the summary statistics for the variable `imdb_score`. The summary statistics provided are the two measures of center (mean and median) and two measures of spread (standard deviation and the quartile values to calculate the IQR) for IMDb score. 

* Highlight and run lines 1 -- 9 in the provided `R` script file to load the data set.  Check that the summary statistics match the output given in the coursepack.

```{r, echo=TRUE, collapse = FALSE}
# Read in data set
movies <- read.csv("https://math.montana.edu/courses/s216/data/Movies2016_F24.csv") 
movies %>% # Data set piped into...
  summarise(favstats(imdb_score)) # Apply favstats function to imdb_score
```

1. Report the values for the two measures of center (mean and median).

\vspace{0.5in}

2. Calculate the interquartile range (IQR = Q3 $-$ Q1) of IMDb scores.

\vspace{0.5in}

3. Report the value of the standard deviation and interpret this value in context of the problem.
\vspace{0.8in}

#### Displaying a single quantitative variable {-}

There are three type of plots used to plot a single quantitative variable: a dotplot, a histogram or a boxplot.  A dotplot of IMDb scores would plot a dot for the IMDb score for each movie released in 2016.

We will create both a histogram and a boxplot of the variable `IMDb`.  

* Enter the name of the variable in both line 16 and line 23 for `variable` in the R script file.  

* Replace the word title for each plot (lines 18 and 25) between the quotations with a descriptive title.  **A title should include: type of plot, variable or variables plotted, and observational units.**  

* Highlight and run lines 15 -- 27 to create a histogram and boxplot. 

Notice that the **bin width** for the histogram is 0.5.  For example the first bin consists of the number of movies in the data set with an IMDb score of 3.25 to 3.75.  It is important to note that a movie with a IMDb score on the boundary of a bin will fall into the bin above it; for example, 4.75 would be counted in the bin 4.75--5.25.  

```{r, out.width="60%", echo=TRUE, eval=FALSE}
movies %>% # Data set piped into...
ggplot(aes(x = variable)) +   # Name variable to plot
  geom_histogram(binwidth = 0.5) +  # Create histogram with specified binwidth
  labs(title = "Title", # Title for plot
       x = "IMDb Score", # Label for x axis
       y = "Frequency") # Label for y axis

movies %>% # Data set piped into...
ggplot(aes(x = variable)) +   # Name variable to plot
  geom_boxplot() +  # Create histogram with specified binwidth
  labs(title = "Title", # Title for plot
       x = "IMDb Score", # Label for x axis
       y = "") # Remove y axis label

```

4. Click the left arrow button on the Plots tab in the lower right-hand corner of the RStudio screen to toggle back to the histogram.  What is the shape of the distribution of IMDb scores?

\vspace{0.2in}


5. Which range of IMDb scores have the highest frequency?

\vspace{0.2in}

6. Click on the right arrow bottom on the Plots tab in the lower right-hand corner of the RStudio screen to toggle back to the boxplot.  Sketch the boxplot created and identify the values of the 5-number summary (minimum value, Q1, median, Q3, maximum value) on the plot.  Use the following formulas to find the invisible fence on both ends of the distribution.  Draw a dotted line at the invisible fence to show how the outliers were found.

$$\text{Lower Fence: values} \le \text{Q}1 - 1.5\times\text{IQR}$$

$$\text{Upper Fence: values} \ge \text{Q}3 + 1.5\times\text{IQR}$$
\vspace{1.8in}

#### Robust Statistics {-}

Let’s examine how the presence of outliers affect the values of center and spread. For this part of the activity we will look at the variable revenue in the movies data set.

```{r, echo=TRUE, eval=TRUE}
movies %>% # Data set piped into...
    summarise(favstats(revenue_mil))

movies %>% # Data set piped into...
    ggplot(aes(x = revenue_mil)) + # Name variable to plot
    geom_boxplot() + # Create histogram with specified binwidth
    labs(title = "Histogram of Revenue of Movies in 2016", # Title for plot
         x = "Revenue (in Millions)", # Label for x axis
         y = "Frequency") # Label for y axis
```

7. Report the two measures of center for these data.

\vspace{0.5in}

8. Report the two measures of spread for these data.

\vspace{0.5in}

To show the effect of outliers on the measures of center and spread, the largest values in the data set were reduced by 100 $MM. This variable is called `revenue_reduced`

```{r, echo=TRUE, eval=TRUE}
movies %>% # Data set piped into...
    summarise(favstats(revenue_reduced))

movies %>% # Data set piped into...
ggplot(aes(x = revenue_reduced)) + # Name variable to plot
geom_boxplot() + # Create histogram with specified binwidth
labs(title = "Histogram of Decreased Revenue of Movies in 2016", # Title for plot
x = "Revenue (in Millions)", # Label for x axis
y = "Frequency") # Label for y axis
```

9. Report the two measures of center for this new data set.
\vspace{0.5in}

10. Report the two measures of spread for this new data set.
\vspace{0.5in}

11. Which measure of center is robust to outliers? Explain your answer.
\vspace{0.8in}

12. Which measure of spread is robust to outliers? Explain your answer.
\vspace{0.8in}

#### Summary statistics for a single categorical and single quantitative Variable{-}

Is there an association between content rating and budget for movies released in 2016?  To use the `favstats()` function in the mosaic package with two variables, we will enter the variables as a formula, response~explanatory.  This function will give the summary statistics for budget for each content rating.

* Highlight and run lines 31--33 in the provided `R` script file and check that the summary statistics match those provided in the coursepack.

```{r, echo=TRUE, collapse=FALSE}
movies %>% # Data set piped into...
  filter(content_rating != "Not Rated") %>% # Remove Not Rated movies
  reframe(favstats(budget_mil~content_rating)) # Find the summary measures for each content rating
```

13.  Which content rating has the largest IQR?

\vspace{0.8in}

14.  Report the mean budget amount for the PG rating.  Use appropriate notation.

\vspace{0.3in}

15.  Report the mean budget amount for the R rating.  Use appropriate notation.

\vspace{0.3in}

16. Calculate the difference in mean budget amount for movies released in 2016 with a PG rating minus those with a R rating.  Use appropriate notation with informative subscripts.

\vspace{0.8in}

17. Interpret the difference in means calculated in question 10 in context of the problem.

\vspace{0.5in}


#### Displaying a single categorical and single quantitative variable {-}

The boxplot of movie budgets (in millions) by content rating is plotted using the code below.  

* Enter the variable `budget_mil` for `response` and the variable `content_rating` for explanatory at line 40.

* Highlight and run code lines 38--44. This plot compares the budget for different levels of content rating.

```{r, out.width="60%", echo=TRUE, eval=FALSE}
movies %>%  # Data set piped into...
  filter(content_rating != "Not Rated") %>% # Remove Not Rated movies
  ggplot(aes(y = response, x = explanatory))+  # Identify variables
  geom_boxplot()+  # Tell it to make a box plot
  labs(title = "Side-by-side Box Plot of Budget by Content Rating for Movies Released in 2016",  
       # Title
       x = "Content Rating",    # x-axis label
       y = "Budget (in Millions)")  # y-axis label
```

18. Answer the following questions about the box plots created.

   a. Which content rating has the highest center?
\vspace{0.2in}

   b. Which content rating has the largest spread?
\vspace{0.2in}

   c. Which content rating has the most skewed distribution?
\vspace{0.2in}

   d. Fifty percent of movies released in 2016 with a PG-13 content rating fall below what value?  What is the name of this value?
\vspace{0.4in}

   e.  What is the value for the first quartile (Q1) for the PG-13 rating?  Interpret this value in context.
\vspace{.8in}

### Take-home messages

1.	Histograms, box plots, and dot plots can all be used to graphically display a single quantitative variable.  

2.  The box plot is created using the five number summary: minimum value, quartile 1, median, quartile 3, and maximum value.  Values in the data set that are less than $\text{Q}_1 - 1.5\times \text{IQR}$ and greater than $\text{Q}_3 + 1.5\times \text{IQR}$ are considered outliers and are graphically represented by a dot outside of the whiskers on the box plot.

3.  Data should be summarized numerically and displayed graphically to give us information about the study.

4.  When comparing distributions of quantitative variables we look at the shape, center, spread, and for outliers.  There are two measures of center: mean and the median and two measures of spread: standard deviation and the interquartile range, IQR = Q3 $-$ Q1. 


### Additional notes

Use this space to summarize your thoughts and take additional notes on today's activity and material covered.

\newpage