---
output:
  pdf_document: default
  html_document: default
---
## Activity 9:  Impacts on the P-value and Confidence Interval

\setstretch{1}

```{r, echo=FALSE}
set.seed(216)
```


### Learning outcomes

* Explore impact of sample size, direction of the alternative hypothesis, and value of the sample statistic on the p-value.

* Explore impact of sample size and confidence level on the width of the confidence interval.


### Terminology review

In today's activity, we will cover how different parameters may impact the p-value and confidence interval. Some terms covered in this activity are:

* Impacts of the sample statistic

* Impacts of sample size

* Effect of confidence level


#### Impacts on the p-value {-}

Let's see how each of the following with change the p-value.  We will revisit the male professional boxer study. In this study, the researchers found that 81 out of 500 male professional boxers were left-handed.  The following is the null distribution with 10,000 simulations assuming that the true proportion of male professional boxers that are left-handed is 0.1.

```{r, echo=TRUE, out.width="60%"}
one_proportion_test(probability_success=0.1,
                    sample_size=500,
                    number_repetitions=10000,
                    as_extreme_as=0.162,
                    direction="greater",
                    summary_measure="proportion")
```

\newpage

In another sample of 500 male professional boxers, only 64 selected the helper toy.  The null distribution created below reflects this new sample.

```{r, echo=TRUE, eval = TRUE}
one_proportion_test(probability_success = 0.1, # Null hypothesis value
          sample_size = 500, # Enter sample size
          number_repetitions = 10000, # Enter number of simulations
          as_extreme_as = 0.128, # Observed statistic
          direction = "greater", # Specify direction of alternative hypothesis
          summary_measure = "proportion") # Reporting proportion or number of successes?
```
1.  Note that this statistic ($\frac{64}{500} = 0.128$) is closer to the null value of 0.1 than the original statistic ($\frac{81}{500}=0.1645$). Is the p-value for this new study larger or smaller than for the original statistic?

\vspace{0.3in}

In activity 7, the following standardized proportion was calculated.

$$Z = \frac{0.164-0.1}{0.013} = 4.923$$

For this new study, the standard error for the sample proportion is the same.

$$SE_0(\hat{p})=\sqrt{\frac{0.1(1-0.1)}{500}} = 0.013$$

2. Calculate the standardized statistic for the new study using a sample proportion of 0.128 ($\hat{p} = 0.128$). 
\vspace{0.5in}

\newpage

3. Illustrate how you would find the p-value from the standardized statistic from this new study on the following standard normal distribution.

```{r Norm-2, warning=FALSE, out.width="50%", echo = FALSE, fig.align='center', fig.cap="Standard Normal Distribution"}
openintro::normTail()
```

   Would this p-value be smaller or larger than the one found in Activity 7?
   
\vspace{0.2in}

#### Change in sample size {-}

In still another study, the researchers took a sample of 300 infants and found the same proportion ($\hat{p}=0.162$) of male boxers that are left-handed.  The null distribution created below reflects this new study.

```{r, echo=TRUE, eval = TRUE}
one_proportion_test(probability_success = 0.1, # Null hypothesis value
          sample_size = 300, # Enter sample size
          number_repetitions = 10000, # Enter number of simulations
          as_extreme_as = 0.162, # Observed statistic
          direction = "greater", # Specify direction of alternative hypothesis
          summary_measure = "proportion") # Reporting proportion or number of successes?
```
4.  Is the p-value found for the smaller sample size ($n=300$) larger or smaller than the p-value found with the larger sample size ($n=500$).

\vspace{0.3in}

5.  Summarize how each of the following affected the p-value: 

a) Using a larger sample size.

\vspace{0.4in}

b) Using a sample statistic closer to the null value.

\vspace{0.4in}


#### Effect of Confidence Level {-}

To look at the impact on confidence interval, we will again look at the male professional boxer study.  The following bootstrap distribution shows a 95\% confidence interval for the parameter of interest.

```{r, echo=TRUE, out.width="60%"}
one_proportion_bootstrap_CI(sample_size = 500,
                            number_successes = 81,
                            number_repetitions = 10000,
                            confidence_level = 0.95)
```

6.  Suppose instead of finding a 95\% confidence interval, we found a 90\% confidence interval.  Would you expect the 90\% confidence interval to be narrower or wider?  Explain your answer.

\vspace{0.4in}

7.  The following R code produced the bootstrap distribution with 10000 simulations that follows.  Circle the value that changed in the code. 

```{r, echo=TRUE, out.width="70%"}
one_proportion_bootstrap_CI(sample_size = 500, # Sample size
                    number_successes = 81, # Observed number of successes
                    number_repetitions = 10000, # Number of bootstrap samples to use
                    confidence_level = 0.90) # Confidence level as a decimal
```


8.  Report both the 95\% confidence interval and the 90\% confidence interval (question 7).  Is the 90\% confidence interval narrower or wider than the 95\% confidence interval?

\vspace{0.5in}

#### Effect of sample size on the width of the confidence interval {-}

How would an decrease in sample size impact the width of the confidence interval?  We will again use a sample size of 300 male boxers that shows the same proportion ($\hat{p}=0.162$) of male boxers that are left-handed.  

The standard error of the sample proportion for this study with the smaller sample size is:

$$SE(\hat{p}) = \sqrt{\frac{0.162\times (1-0.162)}{300}} = 0.0213$$
 
9.  Is the standard error of the sample proportion for this study smaller or larger than the value calculated earlier?

\vspace{0.3in}
 
Recall that the $z^*$ multiplier is 1.96 for a 95\% confidence interval.

10. Calculate the 95\% confidence interval for this study with the smaller sample size.

\vspace{1in}

The width of the confidence interval is found by calculating the difference between the upper value and the lower value.  

$$\text{width of CI = upper CI value - lower CI value}$$

\newpage

The center of the confidence interval is the sample statistic and can also be found using the midpoint of the interval.  

$$\text{midpoint} = \frac{\text{lower CI value + upper CI value}}{2}$$

11. Compare the interval found in question 10 to the interval calculated in Activity 8 (prior to question 4).  

* Calculate the center of the interval with the smaller sample size.

\vspace{0.5in}

* Calculate the center of the interval from Activity 8.

\vspace{0.5in}


* Did the center of the interval change?  
    
\vspace{0.2in}
    
* Calculate the width of the interval with the smaller sample size.
    
\vspace{0.5in}
    
    
* Calculate the width of the interval from Activity 8.
    
\vspace{0.5in}
    
    
* Which interval is wider?
    
\vspace{0.4in}
  
The margin of error represents half the width of the confidence interval since we add and subtract the margin of error to the value of the sample statistic.



$$\text{width of CI} = 2 \times \text{ME}$$


    
12.  Using the width of the interval with the smaller sample size calculated in question 11, calculate the margin of error.

\vspace{0.4in}

13.  What impact does decreasing the sample size have on the width of the confidence interval?

\vspace{0.4in}

\newpage

### Take-home messages

1.  The larger the sample size, the smaller the sample to sample variability.  This will result in a larger standardized statistic and more evidence against the null hypothesis.

2.  The farther the statistic is from the null value, the larger the standardized statistic.  This will result in a smaller p-value and more evidence against the null hypothesis.

3. Decreasing the sample size, increases the sample to sample variability, the standard error resulting in a wider confidence interval.


### Additional notes

Use this space to summarize your thoughts and take additional notes on today's activity and material covered.

\newpage