## Activity 12: Hypothesis Testing of a Single Quantitative Variable

\setstretch{1}

### Learning outcomes

* Given a research question involving one quantitative variable, construct the null and alternative hypotheses
in words and using appropriate statistical symbols.

* Investigate the process of creating a null distribution for one quantitative variable

* Find, evaluate, and interpret a p-value from the null distribution

* Assess the conditions to use the t-distribution model for single mean.

* Calculate and interpret the standardized sample mean

* Use the t-distribution to find the p-value for the test


### Terminology review

In today's activity, we will simulation and theory-based methods to analyze a single quantitative variable.  Some terms covered in this activity are:

* Null hypothesis

* Alternative hypothesis

* t-distribution

* Degrees of freedom
  

To review these concepts, see Chapter .... in the textbook.

### College student sleep habits

According to the National Institutes of Health (NIH) (*add reference*), experts recommend adults (>18) get at least 7 hours of sleep per night. A survey was sent to students in four sections of Stat 216 asking about their sleep habits.  Is there evidence that college students get less than the recommended 7 hours of sleep per night, on average?

#### Summarizing quantitative variables {-}

* Download the R script file and data file for this activity

* Upload both files to the RStudio server and open the R script file

* Enter the name of the dataset for datasetname.csv 

* Highlight and run lines 



The `favstats()` function from the `mosaic` package gives the summary statistics for a quantitative variable. The `R` output below provides the summary statistics for the variable `Graduation_Rate`. The summary statistics provided are the two measures of center (mean and median) and two measures of spread (standard deviation and the quartile values to calculate the IQR) for IMDb score. 

* Highlight and run lines 1 -- 12 in the provided `R` script file to load the data set.  Check that the summary statistics match the output given in the coursepack.

* Notice that the 2-year institutions were removed so the observational units for this study are **4-year higher education institutions.**

```{r, echo=TRUE, eval=TRUE, collapse=FALSE}
IPEDS <- read.csv("https://www.math.montana.edu/courses/s216/data/IPEDS_2018.csv") 
IPEDS <- IPEDS %>%
  filter(Sector != "Public 2-year") # Filters the data set to remove Public 2-year
IPEDS <- IPEDS %>%
  filter(Sector != "Private 2-year") # Filters the data set to remove Private 2-year
IPEDS %>%
    summarize(favstats(Graduation_Rate))
```

2. Report the values for the two measures of center (mean and median).

\vspace{0.5in}


3. Calculate the interquartile range (IQR = Q3 $-$ Q1) of Graduation Rates.

\vspace{0.5in}

4. Report the value of the standard deviation and interpret this value in context of the problem.
\vspace{0.8in}

5. Interpret the value of $Q_3$ in context of the study.

\vspace{0.8in}

#### Displaying a single quantitative variable {-}

There are three type of plots used to plot a single quantitative variable: a dotplot, a histogram or a boxplot.  A dotplot of graduation rate would plot a dot for the graduation rate for each 4-year US higher education institution.

First, let's create a histogram of the variable `Graduation_Rate`.

* Enter the name of the variable in line 19 for `variable` in the R script file.  

* Replace the word title for the plot in line 21 between the quotations with a descriptive title.  **A title should include: type of plot, variable or variables plotted, and observational units.**  

* Highlight and run lines 18 -- ?? to create the histogram. 

```{r, out.width="60%", echo=TRUE, eval=FALSE, warning=FALSE}
IPEDS %>% # Data set piped into...
ggplot(aes(x = xx)) +   # Name variable to plot
  geom_histogram(binwidth = 10) +  # Create histogram with specified binwidth
  labs(title = "Don't forget to title the plot!", # Title for plot
       x = "Graduation Rate", # Label for x axis
       y = "Frequency") # Label for y axis

```

Notice that the **bin width** for the histogram is 10.  For example the first bin consists of the number of institutions in the data set with a graduation rate of 0 to 10\%.  It is important to note that a graduation rate on the boundary of a bin will fall into the bin above it; for example, 20 would be counted in the bin 20--30. 

6. Which range of Graduation Rates have the highest frequency?

\vspace{0.2in}

Next we will create a boxplot of the variable `Graduation_Rate`.  

* Enter the name of the variable in line 19 for `variable` in the R script file.  

* Highlight and run lines....

```{r, echo=TRUE, eval=FALSE, warning=FALSE}
IPEDS %>% # Data set piped into...
ggplot(aes(x = variable)) +   # Name variable to plot
  geom_boxplot() +  # Create boxplot with specified binwidth
  labs(title = "Boxplot of Graduation Rates for 4-year Higher Education Institutions", # Title for plot
       x = "Graduation_Rate", # Label for x axis
       y = "") + # Remove y axis label
    theme(axis.text.y = element_blank(), 
          axis.ticks.y = element_blank()) # Removes y-axis ticks
```

7. Sketch the boxplot created and identify the values of the 5-number summary (minimum value, Q1, median, Q3, maximum value) on the plot.  Use the following formulas to find the invisible fence on both ends of the distribution.  Draw a dotted line at the invisible fence to show how the outliers were found.

$$\text{Lower Fence: values} \le \text{Q}1 - 1.5\times\text{IQR}$$

$$\text{Upper Fence: values} \ge \text{Q}3 + 1.5\times\text{IQR}$$
\vspace{1.8in}

When describing plots of quantitative variables we discuss the shape (symmetric or skewed), the center (mean or median), spread (standard deviation or IQR), and if there are outliers present.

8.  What is the shape of the distribution of graduation rates?

\vspace{0.4in}

9.  From which plot (histogram or boxplot) is it easier to determine the shape of the distribution?

\vspace{0.3in}

10. From which plot is it easier to determine if there are outliers?

\vspace{0.3in}

#### Robust Statistics {-}

Let’s examine how the presence of outliers affect the values of center and spread. For this part of the activity we will look at the variable retention rate in the IPEDS data set.  

```{r, echo=TRUE, eval=TRUE}
IPEDS %>% # Data set piped into...
    summarise(favstats(Retention))

IPEDS %>% # Data set piped into...
    ggplot(aes(x = Retention)) + # Name variable to plot
    geom_boxplot() + # Create boxplot 
    labs(title = "Boxplot of Retention Rates for 4-year Higher Education Institutions", # Title for plot
         x = "Retention Rates (%)", # Label for x axis
         y = "Frequency") # Label for y axis
```

11. Report the two measures of center for these data.

\vspace{0.5in}

12. Report the two measures of spread for these data.

\vspace{0.5in}

To show the effect of outliers on the measures of center and spread, the smallest values of retention rate in the
data set were increased by 30%. This variable is called `Retention_Inc`.

```{r, echo=TRUE, eval=TRUE}
IPEDS %>% # Data set piped into...
    summarise(favstats(Retention_Inc))

IPEDS %>% # Data set piped into...
    ggplot(aes(x = Retention_Inc)) + # Name variable to plot
    geom_boxplot() + # Create histogram 
labs(title = "Boxplot of Increased Retention Rates for 4-year Higher Education Institutions", # Title for plot
x = "Retention Rates (%)", # Label for x axis
y = "Frequency") # Label for y axis
```

13. Report the two measures of center for this new data set.
\vspace{0.5in}

14. Report the two measures of spread for this new data set.
\vspace{0.5in}

15. Which measure of center is robust to outliers? Explain your answer.
\vspace{0.8in}

16. Which measure of spread is robust to outliers? Explain your answer.
\vspace{0.8in}

### Take-home messages

1.	Histograms, box plots, and dot plots can all be used to graphically display a single quantitative variable.  

2.  The box plot is created using the five number summary: minimum value, quartile 1, median, quartile 3, and maximum value.  Values in the data set that are less than $\text{Q}_1 - 1.5\times \text{IQR}$ and greater than $\text{Q}_3 + 1.5\times \text{IQR}$ are considered outliers and are graphically represented by a dot outside of the whiskers on the box plot.

3.  Data should be summarized numerically and displayed graphically to give us information about the study.

4.  When comparing distributions of quantitative variables we look at the shape, center, spread, and for outliers.  There are two measures of center: mean and the median and two measures of spread: standard deviation and the interquartile range, IQR = Q3 $-$ Q1. 


### Additional notes

Use this space to summarize your thoughts and take additional notes on today's activity and material covered.

\newpage