## Activity 11: Inference for a Single Quantitative Variable: Simulation Methods

\setstretch{1}

### Learning outcomes

* Given a research question involving one quantitative variable, construct the null and alternative hypotheses
in words and using appropriate statistical symbols.

* Investigate the process of creating a null distribution for one quantitative variable.

* Find, evaluate, and interpret a p-value from the null distribution.

### Terminology review

In today's activity, we will use simulation-based to analyze a single quantitative variable.  Some terms covered in this activity are:

* Null hypothesis

* Alternative hypothesis

* Null distribution

* Bootstrap distribution

* p-value

To review these concepts, see Chapters 9 and 17 in the textbook.

### Body Temperature

It has long been reported that the mean body temperature of adults is $98.6^{\circ}$F.  There have been a few articles that challenge this assertion. [@luetkemeier2017]  In 2018, a sample of 52 Stat 216 undergraduates were asked to report their body temperature.  Is there evidence that the average body temperature of Stat 216 undergraduates differs from the known temperature of $98.6^{\circ}$F?

\vspace{2mm}
* Observational units:

\vspace{2mm}
* Variable:

    * Units?:

\vspace{2mm}
#### R Instructions {-}

* Download the R script file and data file for this activity

* Upload both files to the RStudio server and open the R script file

* Enter the name of the dataset for datasetname.csv in line 8

* Highlight and run lines 1--8 to load the data set

```{r, echo=TRUE, eval=FALSE}
bodytemp <- read.csv("datasetname.csv")
```

```{r, echo=FALSE, eval=TRUE}
set.seed(216)
bodytemp <- read.csv("https://www.math.montana.edu/courses/s216/data/normal_temperature.csv")
```

\newpage

#### Hypothesis Testing for a Single Mean {-}

**Parameter of interest in context of the study:**

\vspace{0.8in}

**Null Hypothesis (in words):**

\vspace{0.8in}

**Null Hypothesis (in notation):**

\vspace{0.3in}

**Alternative Hypothesis (in words):**

\vspace{0.8in}

**Alternative Hypothesis (in notation):**

\vspace{0.3in}

#### Summarize and visualize the data {-}

The `favstats()` function from the `mosaic` package gives the summary statistics for a quantitative variable.

* Enter the variable name, `Temp`, for `variable` in line 13

* Highlight and run lines 12--13

```{r, echo=TRUE, eval=FALSE, collapse=FALSE}
bodytemp %>%
    summarize(favstats(variable))
```
1. About how far is each body temperature for a Stat 216 undergraduate from the mean body temperature, on average? Report the value and the appropriate notation

\vspace{0.3in}

Create a boxplot of the variable `Temp`.

* Enter the name of the variable in line 19 for `variable` in the R script file. 

* Enter a title in line 21 for the plot between the quotations.

* Highlight and run lines 18--25.

```{r, echo=TRUE, eval=FALSE}
bodytemp %>%
  ggplot(aes(x = variable))+
  geom_boxplot()+
  labs(title="Be sure to include a title!",
       x = "body temperature (*F)", # Label for x axis
       y = "") + # Remove y axis label
        theme(axis.text.y = element_blank(), 
          axis.ticks.y = element_blank()) # Removes y-axis ticks
```


2. Describe the distribution of body temperature using the four characteristics of boxplots.

\vspace{1in}


### Simulation methods {-}

#### Notes on simulation methods for a single mean {-}

\vspace{2in}

To simulate the null distribution of sample means we will use a bootstrapping method.  Recall that the null distribution must be created under the assumption that the null hypothesis is true.  Therefore, before bootstrapping, we will need to *shift* each data point by the difference $\mu_0 - \bar{y}$.  This will ensure that the mean of the shifted data is $\mu_0$ (rather than the mean of the original data, $\bar{y}$), and that the simulated null distribution will be centered at the null value.  

* Calculate the difference $\mu_0 - \bar{y}$.  Based on the sign of this difference, will we need to shift the data up or down?

\vspace{0.4in}

Your instructor will demonstrate how the shift is performed in Excel.

* Open the data set in Excel. 

* Create a new column labeled Shift.

* In the column, Shift, add the shifted value to each value in the `Temp` column.

```{r, echo=TRUE, eval=FALSE, collapse=FALSE}
bodytemp <- read.csv("normal_temperature.csv")
bodytemp %>%
    summarize(favstats(Shift))
```
3. Report the mean of the `Shift` variable.  Why does it make sense that this value is the same as the null value?

\vspace{0.6in}

4. Report the standard deviation of the `Shift` variable. How does this compare to the standard deviation for the variable `Temp`?  Explain why these values are the same.

\vspace{0.6in}

5. What inputs should be entered for each of the following to create the simulated null distribution?
\vspace{1mm}

* Null value (What is the null value for the study?):

\vspace{.15in}
* Summary measure ("mean" or "median"):

\vspace{0.15in}
* Shift (difference between $\mu_0 -\bar{y}$):

\vspace{0.15in}

* As extreme as (enter the value for the observed sample mean):

\vspace{.15in}
* Direction (`"greater"`, `"less"`, or `"two-sided"`):

\vspace{.15in}

* Number of repetitions:

\vspace{.15in}

The `one_mean_test` will be used to find the p-value for the simulation test.  Following the instructions below to complete the code.

* Enter your answers for question 5 in place of the `xx`'s to produce the null distribution with 10000 simulations.

* Highlight and run lines 29--35.

```{r, echo=TRUE, eval=TRUE}
one_mean_test(bodytemp$Temp,#Enter the object name and variable
              null_value = 98.6,
              summary_measure = "mean",  #Can choose between mean or median
              shift = 0.3154, #Difference between the null value and the sample mean
              as_extreme_as = 98.2846, #Value of the summary statistic
              direction = "two-sided", #Specify direction of alternative hypothesis
              number_repetitions = 10000)
```

#### Notes on the null distribution {-}

\vspace{2in}

**Interpretation of the p-value in context of the problem.**

\vspace{1in}

**Conclusion of the test:**

\vspace{1in}


### Simulation methods to create a confidence interval {-}

Unlike creation of the null distribution, the bootstrap distribution we use for creating a confidence interval is found by sampling with replacement from the original sample.  To create one dot on the bootstrap distribution:

* Write the original values for the variable on $n$ cards; one card for each observational unit.

* Sample with replacement from the cards $n$ times.

* Plot the mean from each resample on the bootstrap distribution.

Use the provided R script file to find a 95\% confidence interval.

* Enter the appropriate confidence level for `xx`.

* Highlight and run lines 39--42.

```{r, eval=TRUE, echo=TRUE}
one_mean_CI(bodytemp$Temp, #Enter the name of the variable
            summary_measure = "mean", #choose the mean or median
            number_repetitions = 10000,  # Number of simulations
            confidence_level = 0.95) #Enter as a decimal
```


#### Notes on the bootstrap distribution {-}

\vspace{2in}

**Confidence Interval:**

\vspace{0.2in}

**Interpretation of the confidence interval:**

\vspace{1in}

\newpage

### Take-home messages

1.	We use bootstrapping---sampling with replacement---from the shifted data to generate a null distribution of simulated sample means. In order to ensure that the null distribution is centered at the null value, $\mu_0$, we shift the data by adding $\mu_0 - \bar{y}$ to each value in the original data set. Note that if this value of the shift is negative, we are shifting the data down; if it is positive, we shift the data up.

2.  The mean of the shifted data will equal the null value, $\mu_0$, but the standard deviation of the shifted data will be the same as the standard deviation of the original data.

3.  As in the one proportion scenario, we calculate the p-value for a simulation-based hypothesis test for a single mean by finding the proportion of simulated sample means that are as or more extreme (in the direction of $H_A$) as the observed sample mean, $\bar{y}$.


### Additional notes

Use this space to summarize your thoughts and take additional notes on today's activity and material covered.

\newpage