# Inference for a Single Quantitative Variable

## Vocabulary Review and Key Topics

Review the Golden Ticket posted in the resources at the end of the coursepack for a summary of a single quantitative variable.  Module 6 will cover hypothesis testing using both simulation and theory-based methods.

* The **summary measure** for one quantitative variable is the **mean**

* Additionally, we can find the five number summary (min, Q1, median, Q3, max) as well as the sample standard deviation

*R code to find the summary statistics for a quantitative variable

```{r, eval=FALSE}
object %>% # Data set piped into...
    summarise(favstats(variable))
```

* Quartile 1, $Q_1$: value at the 25th percentile; approximately 25\% of data values are at and below the value of $Q_1$

* Quartile 2, $Q_3$: value at the 75th percentile; approximately 75\% of data values are at and below the value of $Q_3$

* Sample standard deviation, $s$: on average, each value in the data set is s units from the mean of the data set

* **Interquartile range**: $IQR = Q_3-Q_1$ 

Types of plot for one quantitative variables

* **Histogram**: sorts a quantitative variable into bins of a certain width 

* R code to create a histogram

```{r}
object %>% # Data set piped into...
    ggplot(aes(x = variable)) +   # Name variable to plot
    geom_histogram(binwidth = 10) +  # Create histogram with specified binwidth
    labs(title = "Don't forget to title the plot!", # Title for plot
         x = "x-axis label", # Label for x axis
         y = "y-axis label") # Label for y axis
```


* **Boxplot**: plots the values of the five number summary and shows any outliers in the data set

* R code to create a boxplot

```{r, echo=FALSE}
object %>% # Data set piped into...
    ggplot(aes(x = variable)) + # Name variable to plot
    geom_boxplot() + # Create boxplot 
    labs(title = "Don't forget to title the plot!", # Title for plot
         x = "x-axis label", # Label for x axis
         y = "y-axis label") # Label for y axis
```

* **Dotplot**: plots each value as a dot along the x-axis

* Four characteristics of boxplots

    - Shape  (symmetric or skewed)
    
    - Center 
    
    - Spread
    
    - Outliers?
    
### Simulation Hypothesis Testing {-}

Hypotheses:

$$H_0: \mu = \mu_0$$
$$H_A: \mu\left\{
\begin{array}{ll}
< \\
\ne \\
< \\
\end{array}
\right\}
\mu_0 $$

* R code to use for **simulation methods** for one quantitative variable to find the p-value, one_mean_test, is shown below. Review the comments (instructions after the #) to see what each should be entered for each line of code.

```{r, eval=FALSE, echo=TRUE}
one_mean_test(object$variable,#Enter the object name and variable
              null_value = xx, #Enter the null value for the study
              summary_measure = "mean",  #Can choose between mean or median
              shift = xx, #Difference between the null value and the sample mean
              as_extreme_as = xx, #Value of the summary statistic
              direction = "xx", #Specify direction of alternative hypothesis
              number_repetitions = 10000)
```

#### Exploratory data analysis {-}

At the end of this module, you should understand how to calculate a summary statistic and plot a single quantitative variable.  

* Notation for a sample mean: $\bar{x}$

* Notation for a population mean: $\mu$

* Types of plots for a single categorical variable:

    - Histogram
    
    - Boxplot
    
    - Dotplot

\newpage
