# Inference for a Single Quantitative Variable

## Vocabulary Review and Key Topics

Review the Golden Ticket posted in the resources at the end of the coursepack for a summary of a single quantitative variable.  Module 7 will cover creating confidence interval using both simulation and theory-based methods. Additionally, we learn about errors and power.


* **Theory-based methods**: when specific conditions are met, a data can be fit with a theoretical distribution

* **Conditions for the sampling distribution of $\bar{x}$ to follow an approximate normal distribution**:

    * **Independence**: The sample’s observations are independent, e.g., are from a simple random sample. (*Remember*: This also must be true to use simulation methods!)

     * **Large enough sample size: Normality Condition**: The sample observations come from a normally distributed population.  To check use the the following rules of thumb:
     
         - $n < 30$: The distribution of the sample must be approximately normal with no outliers
         
         - $30 \ge n < 100$: We can relax the condition a little; the distribution of the sample must have no extreme outliers or skewness
         
         - $n > 100$: Can assume the sampling distribution of $\bar{x}$ is nearly normal, even if the underlying distribuion of individual observationals is not
         
* **t-distribution**: a theoretical distribution that is symmetric with a given degrees of freedom ($n-1$)

    * $t_{n-1}$

* **Standardized sample mean**: standardized statistic for a single quantitative variable calculated using:

$$
T = \frac{\bar{x} - \mu_0}{SE(\bar{x})},
$$

* **Standard error of the sample mean assuming the null is true**: measures the how far each possible sample mean is from the true mean, on average and is calculated using the formula below:

$$SE(\bar{x})=\frac{s}{\sqrt{n}}$$

* The following R code is used to find the p-value using theory based methods for a single quantitative variables.

    * pt will give you a p-value using the t-distribution with n-1 df (enter for yy)
    
    * Enter the value of the standardized statistic for xx

    * If a greater than alternative, change lower.tail = TRUE to FALSE.
    
    * If a two-sided test, multiply by 2.
    
```{r, eval = FALSE}
pt(xx, df = yy, lower.tail=TRUE)
```

* Calculation of the confidence interval for a sample mean

$$\bar{x}\pm t^*\times SE(\bar{x})$$

* R code to find the multiplier for the confidence interval using theory-based methods.

   - qt will give you the multiplier using the t-distribution with $n-1$ df (enter for yy)
    
   - Enter the percentile for the given confidence level

```{r, eval=FALSE, echo=TRUE}
qt(percentile, df=yy, lower.tail=FALSE)
```

\newpage
