## Video Notes: Theory-based Inference for a single quantitative variable 

Read Chapters 5, 17, and 12 in the course textbook.  Use the following videos to complete the video notes for Module 7.

### Course Videos

* Theoretical_Tests_One_Quantitative_Variable

* Theoretical_Intervals_One_Quantitative_Variable

* Errors_Power

\setstretch{1}

### Theory-based Test for One Quantitative Variable (sections 17.3.1, 17.3.2 and 17.3.4) {-}

Conditions for inference using theory-based methods:

- Independence:

\vspace{0.2in}

- Large enough sample size:

\vspace{0.2in}

#### $t$-distribution {-}

In the theoretical approach, we use the Central Limit Theorem (CLT) to tell us that---under certain conditions---the distribution of sample means will be approximately normal, centered at the assumed true mean under $H_0$, and with standard deviation $\frac{\sigma}{\sqrt{n}}$.

$$\bar{y} \sim N\left(\mu_0, \frac{\sigma}{\sqrt{n}}\right)$$
\setstretch{1.5}

* Estimate the population standard deviation, $\sigma$, with the ___________________________ standard deviation, ________.

* For a single quantitative variable we use the ____ - distribution with _______________
degrees of freedom to approximate the sampling distribution.

\setstretch{1}

Equation for the standard error of the sample mean:

\vspace{0.5in}

Equation for the standardized sample mean:

\vspace{0.5in}


General steps of a hypothesis test

1.	Write a research question and hypotheses.

2.	Collect data and calculate a summary statistic.

3.	Model a sampling distribution which assumes the null hypothesis is true.

\rgi \rgi - If the conditions are met, the sampling distribution of a sample mean which assumes the null hypothesis is true is the $t$-distribution with $n-1$ degrees of freedom

4.	Calculate a p-value.

\rgi \rgi - Calculate the standardized sample mean ($T$) and compare that to the $t$-distribution with $n-1$ degrees of freedom.
\rgi \rgi - P-value will be the area under the curve beyond $T$ (in the direction of the alternative hypothesis)

5.	Draw conclusions based on a p-value.


#### Optional notes: Additional Example {-}

Adult male polar bears historically have weighed, on average, 370kg.  Scientists are concerned that reduction in food availability and a shortened hunting season may be negatively impacting their weights. The weight was measured on a representative sample of 83 male polar bears from the Southern Beaufort Sea. Is there evidence that male polar bears in the Southern Beaufort Sea weigh less than 370kg, on average?  

```{r, echo=TRUE}
pb <- read.csv("https://math.montana.edu/courses/s216/data/polarbear.csv")
```

Plots of the data:

```{r, echo=TRUE, out.width="60%"}
pb %>%
    ggplot(aes(x = Weight)) +   # Name variable to plot
    geom_histogram(binwidth = 10) +  # Create histogram with specified binwidth
    labs(title = "Histogram of Male Polar Bear Weight", # Title for plot
       x = "Weight (kg)", # Label for x axis
       y = "Frequency") # Label for y axis

pb %>% # Data set piped into...
ggplot(aes(x = Weight)) +   # Name variable to plot
  geom_boxplot() +  # Create boxplot
  labs(title = "Boxplot of Male Polar Bear Weight", # Title for plot
       x = "Weight (kg)", # Label for x axis
       y = "") + # Label for y axis  
    theme(axis.text.y = element_blank(), 
          axis.ticks.y = element_blank()) # Removes y-axis ticks

```

Summary Statistics:

```{r, echo=TRUE}
pb %>%
  summarise(favstats(Weight)) #Gives the summary statistics
```

Calculate the standard error of the sample mean weight:

\vspace{0.3in}

Calculate the standardized sample mean weight (the $T$-score):

\vspace{0.5in}

Interpret the standardized statistic:

\vspace{0.8in}

Label the standardized sample mean (standardized statistic) on the $t$-distribution with 82 degrees of freedom below and shade the area representing the p-value.  Why 82 degrees of freedom?

\vspace{0.3in}

```{r, pvaluepb, echo = F}

x <- seq(-4, 4, length=100)
hx<-dt(x, 82)
degf <- 82

plot(x, hx, type="l", lty=1, lwd=3, xlab="",
  ylab="Density", main="t-distribution with 82 df")

```


To find the theory-based p-value:

```{r, echo=TRUE}
pt(-4.683, df=82, lower.tail=TRUE)
```

Interpretation of the p-value:

* Statement about probability or proportion of samples

* Statistic (summary measure and value) and Direction of the alternative 
    
* Null hypothesis (population reference, summary measure, equal to null value)

* Context of the problem (observational units, variable (if variable is quantitative, include units))

\vspace{0.6in}

Conclusion:

* Amount of evidence
    
* For the alternative hypothesis (population reference, summary measure, direction, null value)

* Context (observational units, variable (if variable is quantitative, include units))

\vspace{0.5in}

Decision at a significance level of 0.05 $(\alpha = 0.05)$:

\vspace{0.3in}

Generalization:

* Can the results of the study be generalized to the target population?

\vspace{0.4in}


### Theory-based Confidence Interval for One Quantitative Variable (section 17.3.3) {-}

Confidence intervals are used to estimate the _________________.

General formula for a confidence interval is always:

\rgi \rgi $\text{statistic} \pm \text{margin of error}$

\rgi \rgi where $\text{margin of error} = \text{multiplier} \times \text{standard error of the statistic}$

For a single quantitative variable,

* $\text{statistic}$ = ___________

* $\text{margin of error} = \text{multiplier} \times \text{standard error of the statistic} = $ ____________ $\times$ ______________

The formula for the standard error of the sample mean is the same as was used in the previous video: $SE(\overline{y}) = \frac{s}{\sqrt{n}}$

The $t^*$ multiplier is the value at the given percentile(s) of the t-distribution with $n - 1$ degrees of freedom.

#### Optional notes: Additional Example {-}


What is the average weight of adult male polar bears?  The weight was measured on a representative sample of 83 male polar bears from the Southern Beaufort Sea.

```{r, echo=TRUE}
pb <- read.csv("https://math.montana.edu/courses/s216/data/polarbear.csv")
```

Plots of the data:

```{r, echo=TRUE, out.width="60%"}
pb %>%
    ggplot(aes(x = Weight)) +   # Name variable to plot
    geom_histogram(binwidth = 10) +  # Create histogram with specified binwidth
    labs(title = "Histogram of Male Polar Bear Weight", # Title for plot
       x = "Weight (kg)", # Label for x axis
       y = "Frequency") # Label for y axis

pb %>% # Data set piped into...
ggplot(aes(x = Weight)) +   # Name variable to plot
  geom_boxplot() +  # Create boxplot
  labs(title = "Boxplot of Male Polar Bear Weight", # Title for plot
       x = "Weight (kg)", # Label for x axis
       y = "") + # Label for y axis  
    theme(axis.text.y = element_blank(), 
          axis.ticks.y = element_blank()) # Removes y-axis ticks

```

Summary Statistics:

```{r, echo=TRUE}
pb %>%
  summarise(favstats(Weight)) #Gives the summary statistics
```

Calculate the standard error of the sample mean weight:

\vspace{0.3in}

To find the $t^*$ multiplier for a 90% confidence interval:

```{r, echo=TRUE}
qt(0.95, df = 82)
```

Draw a line at the provided $t^\star$ on the $t$-distribution with 82 degrees of freedom shown below, and label the percentile of the $t^\star$ value.  

```{r, tstarpb2, echo = F}

x <- seq(-4, 4, length=100)
hx<-dt(x, 82)
degf <- 82

plot(x, hx, type="l", lty=1, lwd=3, xlab="",
  ylab="Density", main="t-distribution with 82 df")

```

Why did we use the 95th percentile of the appropriate $t$-distribution to calculate a 90% confidence interval?

\vspace{0.4in}

Calculate the margin of error for the 90% confidence interval from these data:

\vspace{0.4in}


Calculate the 90% confidence interval for the true mean weight of polar bears from the Southern Beaufort Sea:

\vspace{0.8in}

Confidence interval interpretation:

* How confident you are (e.g., 90%, 95%, 98%, 99%)
    
* Parameter of interest (including context: observational units, variable (if variable is quantitative, include units))
    
* Calculated interval
    
\vspace{0.8in}


### Decisions, Errors, and Power - Video Chapter12 {-}

\setstretch{1.25}

Significance level: arbitrary cut-off set by the researcher for deciding if a p-value is ___________ or _____________.

* Notation:

\vspace{0.2in}

Decision: To either _______________ or _____________________________ the null hypothesis.

* 	If p-value _____ $\alpha$, the p-value is considered “small”

    * 	____________________________________ evidence against the null hypothesis.

    * Decision: ______________ the null hypothesis.
	 
    *	We say the results _______________ statistically significant.
    
* If p-value _____ $\alpha$, the p-value is considered “large”

    * ____________________________________ evidence against the null hypothesis.
    
    * Decision: ________________________________ the null hypothesis.
    
    * We say the results _______________ statistically significant.
    
Errors: 

* Type I error: __________________________ the null hypothesis even though the null hypothesis is __________________.

    * Conclude the ________________ hypothesis is true when really the _______________
    hypothesis is true.
    
    * Anytime we __________________________ the null hypothesis, we could be making a type I error!
    
* Type II error: _____________________________ the null hypothesis even though the null hypothesis is __________________.

    * Conclude the _________________ hypothesis is true when really the ______________
    hypothesis is true.

    * Anytime we _____________________________ the null hypothesis, we could be making a type II error!

* Probability of a type I error = 

Confirmation bias: selecting the sign in the alternative hypothesis based off ______________________.

*	This increases the chance of making a type _________ error.

\setstretch{1}

Example: Polar bears

* $H_0:\mu=370; H_A:\mu <370$; where $\mu$ represents the true mean weight of adult male polar bears in the Southern Beaufort Sea region.

\setstretch{1.25}

* P-value was less than 0.0001

*	Decision at a 5% significance level?  ____________________ the null hypothesis.

    * Possible type ________ error

    * Interpretation of that error: We conclude ________________________________ when really _______________________________________________________.
    
    * Probability that this is a type ______ error? __________

Statistical significance versus practical importance:

* Statistically significant results: If p-value ___________ $\alpha$ 

* Practically important results: If the difference seen is __________________.

*	Small sample sizes tend to have ____________ p-values, so the results may be _______________ but not ______________.

* Large sample sizes tend to have ____________ p-values, so the results may be _______________ but not_______________.


Power

*	Probability of ______________ the null hypothesis IF the null hypothesis is _______________.

*	Impacts on power:

    *	Larger significance level = ______________ power
    
    *	Larger sample size = _____________ power
    
    *	One-sided alternative hypothesis = ______________ power
    
    *	Smaller sample standard deviation = ______________ power
    
    * True value being farther from the null value = ______________ power

\setstretch{1}

### Concept Check

Be prepared for group discussion in the next class. One member from the table should write the answers to the following on the whiteboard.

1. Are the conditions met to analyze the polar bear data using theory-based methods?

\vspace{0.7in}

2. Interpret the confidence interval found with theory-based methods.

\vspace{0.2in}

3. What is the power of the test?

\vspace{0.6in}

 
\newpage