## Module 8 Lab:  Is yawning contagious?

\setstretch{1}

### Learning outcomes

* Given a research question involving two categorical variables, construct the null and alternative hypotheses
  in words and using appropriate statistical symbols.
  
* Assess the conditions to use the normal distribution model for a difference in proportions.

* Describe and perform a simulation-based hypothesis test for a difference in proportions.

* Calculate the Z test statistic for a difference in proportions.

* Find, interpret, and evaluate the p-value for a hypothesis test for a difference in proportions.

* Create and interpret a confidence interval for a difference in proportions.

### Is yawning contagious?

Conventional wisdom says yawns are contagious; when you see someone else yawn, you’re prone to feel sleepy and let out a yawn yourself. How many times have you caught yourself in this situation, or noticed it in someone else? 

The folks at MythBusters, a popular television program on the Discovery Channel, investigated this issue by using a two-way mirror and a hidden camera. Fifty participants who volunteered to be on the show sat in a booth, accompanied only by an experimental attendee. For some of the participants, the attendee yawned (planting a yawn “seed”), while for other participants the attendee did not yawn. The researchers decided in advance, with a random mechanism, which participants would receive the yawn seed and which would not. As time passed, the researchers watched to see which participants yawned.

Is there evidence that planting a "yawn seed" increases the likelihood of a subject to yawn?

* Upload and open the R script file for the Two Categorical Variable Lab. 

* Upload the csv file, `YawnData.csv`.  Enter the name of the data set for `datasetname` in the R script file in line 7. 

* Highlight and run lines 1--8 to get the counts for each combination of categories.

```{r, out.width="60%", echo=TRUE, eval=FALSE}
yawn <- read.csv("datasetname")
yawn %>% group_by(YawnSeed) %>% count(Outcome)
```

1. Is this an experiment or an observational study?
\vspace{0.2in}

2. Complete the following two-way table using the R output.

\begin{center}
\begin{tabular}{|c|c|c|c|}\hline
 & \multicolumn{2}{|c|}{\textbf{Treatment}} & \\ \hline
\textbf{Outcome} & \hspace{0.35in} Yawn Seed \hspace{0.35in} & \hspace{0.35in} No Yawn Seed \hspace{0.35in} & \hspace{0.35in} Total \hspace{0.35in} \\ \hline
 Yawned & & & \\ 
  & & & \\ \hline
 Did Not Yawn & & & \\ 
  & & & \\ \hline
 Total & & &  \\ 
 & & & \\ \hline  
\end{tabular}
\end{center}

3. Write the parameter of interest for the research question.
\vspace{0.6in}

4. Write the null and alternative hypotheses in notation.

\vspace{2mm}

|    $H_0$:

\vspace{0.2in}

|    $H_A$:

\vspace{0.2in}

5. **Calculate the summary statistic (difference in proportions).  Use appropriate notation.**
\vspace{0.4in}

* Fill in the missing values/names in the R script file in the two-proportion_test function to create the null distribution and find the simulation p-value for the test.

```{r, echo=TRUE, eval=FALSE}
two_proportion_test(formula = response~explanatory, # response ~ explanatory
         data= yawn, # Name of data set
         first_in_subtraction = "xx", # Order of subtraction: enter the name of Group 1
         number_repetitions = 1000, # Always use a minimum of 1000 repetitions
         response_value_numerator = "xx", # Define which outcome is a success 
         as_extreme_as = xx, # Calculated observed statistic (difference in sample proportions)
         direction="xx") # Alternative hypothesis direction ("greater","less","two-sided")
```

6. Report the p-value. How much evidence does the p-value provide against the null hypothesis?
\vspace{0.3in}

7. **Calculate the number of standard errors the sample difference in proportion is from the null value of zero.**
\vspace{0.7in}

* Enter the value of Z in the pnorm function in the R script file to find the theory-based p-value.  

```{r, echo=TRUE, eval=FALSE}
pnorm(xx, lower.tail = FALSE)
```

8. Explain why the theory-based p-value does not match the p-value found using simulation methods.

\vspace{0.8in}

#### Confidence Interval {-}

9.  **Will a 95\% simulation confidence interval contain the null value of zero? Explain your answer.**
\vspace{0.7in}

\newpage

* Fill in the missing values/names in the R script file in the two-proportion_bootstrap_CI function to create the null distribution and find the 90\% simulation confidence interval.

```{r, echo=TRUE, eval=FALSE}
two_proportion_bootstrap_CI(formula = response~explantory, # response~explanatory,
                            data = yawn, # Name of data set
                            first_in_subtraction = "xx", # Order of subtraction; enter the name of Group 1
                            number_repetitions = 1000, #Always use a minimum of 1000 repetitions
                            response_value_numerator = "xx", # Define which outcome is a success
                            confidence_level = xx) #Enter the level of confidence as a decimal)
```

10. Report the 90\% confidence interval.

\vspace{0.2in}

11. What type of error may have occurred?

\vspace{0.2in}

12. Write a paragraph summarizing the results of the study.  Be sure to describe:

* Summary statistic and interpretation

    * Summary measure (in context)
    
    * Value of the statistic 
    
    * Order of subtraction when comparing two groups

* P-value and interpretation

    * Statement about probability or proportion of samples
    
    * Statistic (summary measure and value)
    
    * Direction of the alternative 
    
    * Null hypothesis (in context) 


* Confidence interval and interpretation

    * How confident you are (e.g., 90%, 95%, 98%, 99%)
    
    * Parameter of interest
    
    * Calculated interval
    
    * Order of subtraction when comparing two groups


* Conclusion (written to answer the research question)

    * Amount of evidence
    
    * Parameter of interest 
    
    * Direction of the alternative hypothesis


* Scope of inference

    * To what group of observational units do the results apply (target population or observational units similar to the sample)?
	
    * What type of inference is appropriate (causal or non-causal)?

**Upload a copy of your group's paragraph to Gradescope.** 

\newpage

Paragraph (continued): 

\newpage
