## Lecture Notes Week 9: Theoretical Inference for Two Categorical Variables 


\setstretch{1}

### Hypothesis testing using theory-based methods {-}

Conditions for inference using theory-based methods:

- Independence:

\vspace{0.2in}

- Large enough sample size:

\vspace{1in}

* Calculate the standardized statistic

* Find the area under the standard normal distribution at least as extreme as the standardized statistic

Equation for the standard error of the difference in sample proportions assuming the null hypothesis is true:

\vspace{0.5in}

\setstretch{1.5}

* This value measures how far each possible sample difference in _____________ is from the _________ value, on average.

\setstretch{1}

Equation for the standardized differenece in sample proportions:

\vspace{0.5in}

\setstretch{1.5}

* This value measures how many _____________ deviations the sample difference in _____________ is above/below the __________ value.

\setstretch{1}

Example: In the week 3 Lab, we investigated data on higher education institutions in the United States, collected by the Integrated Postsecondary Education Data System (IPEDS) for the National Center for Education Statistics (NCES). A random sample of 2900+ higher education institutions in the United States was collected in 2018.  Two variables measured on this data set is whether the institution is a land grant university and whether the institution offers tenure.  Does the proportion of universities that offer tenure differ between land grant and non-land-grant institutions?

Are the conditions met to analyze the university data using theory-based methods?

\vspace{1in}

```{r, echo=TRUE}
IPED <-read.csv("https://math.montana.edu/courses/s216/data/IPEDS_2018.csv")

IPEDS <- IPED %>%
    drop_na(Tenure)

IPEDS %>% # Data set piped into...
    ggplot(aes(x = LandGrant, fill = Tenure)) +   # This specifies the variables
  geom_bar(stat = "count", position = "fill") +  # Tell it to make a stacked bar plot
  labs(title = "Segmented Bar Plot of Tenure Availability \n Use by Type of Institution",  
       # Make sure to title your plot 
       x = "Land Grant",   # Label the x axis
       y = "") + # Remove y axis label 
    scale_fill_grey()

IPEDS %>% group_by(LandGrant) %>% count(Tenure)
```
What is the explanatory variable?

\vspace{0.2in}

What is the response variable?

\vspace{0.2in}

Write the parameter of interest:

\vspace{0.8in}

Hypotheses:

In notation:  

$H_0:$

\vspace{0.2in}

$H_A:$

\vspace{0.2in}

In words: 

$H_0:$

\vspace{0.6in}

$H_A:$

\vspace{0.6in}

Report the summary statistic:

\vspace{0.8in}

Calculate the standardized difference in sample proportion of higher education institutions that offer tenure between land grant universities and non-land grant universities.

* 1st calculate the standard error of the difference in proportion

\vspace{0.5in}


* Then calculate the Z score.

\vspace{0.5in}

```{r standNormc, warning=FALSE, out.width="50%"}
openintro::normTail(m = 0, s = 1)
```

Interpret the standardized statistic

\vspace{0.5in}

To find the p-value, find the area under the standard normal distribution at the standardized statistic and more extreme.

```{r, echo=TRUE}
pnorm(0.985, lower.tail = FALSE)*2
```
Interpretation of the p-value:

* Statement about probability or proportion of samples

* Statistic (summary measure and value)
    
* Direction of the alternative 
    
* Null hypothesis (in context) 

\vspace{0.8in}

Conclusion with scope of inference: 

* Amount of evidence
    
* Parameter of interest 
    
* Direction of the alternative hypothesis

* Generalization

* Causation

\vspace{0.6in}

### Confidence interval {-}

* Estimate the _______________ in true _______________ 

* $CI = \text{statistic} \pm \text{margin of error}$

\vspace{0.5in}

#### Theory-based method for a two categorical variables {-}

* $CI = \hat{p}_1-\hat{p}_2 \pm (z^* \times SE(\hat{p}_1-\hat{p}_2))$

\setstretch{1.5}

* When creating a confidence interval, we no longer assume the _____________ hypothesis is true. Use the sample _____________ to calculate the sample to sample variability, rather than $\hat{p}_{pooled}$.

\setstretch{1}

Equation for the standard error of the difference in sample proportions *NOT* assuming the null is true:

\vspace{0.5in}

Example:  Estimate the difference in true proportions of higher education institutions that offer tenure between land grant universities and non-land grant universities.

Find a 90\% confidence interval:

* 1st find the $t^*$ multiplier 

```{r, echo=TRUE}
qnorm(0.95, lower.tail=TRUE)
```

* Next, calculate the standard error for the difference in proportions

\vspace{0.8in}

* Calculate the margin of error

\vspace{0.6in}

* Calculate the endpoints of the 90\% confidence interval

\vspace{0.6in}

Confidence interval interpretation:

* How confident you are (e.g., 90%, 95%, 98%, 99%)
    
* Parameter of interest
    
* Calculated interval
    
* Order of subtraction when comparing two groups

\vspace{0.8in}


\newpage
