## Video Notes: Theoretical Inference for Two Categorical Variables 

Read Sections 15.3 and 15.4 in the course textbook.  Use the following videos to complete the video notes for Module 9.

### Course Videos

* 15.3TheoryTests

* 15.3TheoryIntervals

<!-- * RelativeRisk -->


\setstretch{1}

### Hypothesis testing using theory-based methods - Video 15.4TheoryTests {-}

Example: In Modules 3 and 4, we investigated data on higher education institutions in the United States, collected by the Integrated Postsecondary Education Data System (IPEDS) for the National Center for Education Statistics (NCES) [@ipeds]. A random sample of 2900+ higher education institutions in the United States was collected in 2018.  Two variables measured on this data set is whether the institution is a land grant university and whether the institution offers tenure.  Does the proportion of universities that offer tenure differ between land grant and non-land-grant institutions?

What is the explanatory variable?

\vspace{0.2in}

What is the response variable?

\vspace{0.2in}

Write the parameter of interest:

\vspace{0.8in}

Hypotheses:

In notation:  

$H_0:$

\vspace{0.2in}

$H_A:$

\vspace{0.2in}

```{r, echo=TRUE, collapse=FALSE}
IPED <-read.csv("https://math.montana.edu/courses/s216/data/IPEDS_2018.csv")

IPEDS <- IPED %>%
    drop_na(Tenure)

IPEDS %>% # Data set piped into...
    ggplot(aes(x = LandGrant, fill = Tenure)) +   # This specifies the variables
  geom_bar(stat = "count", position = "fill") +  # Tell it to make a stacked bar plot
  labs(title = "Segmented Bar Plot of Tenure Availability 
       by Type of Institution for Higher Ed Institutions",  
       # Make sure to title your plot 
       x = "Land Grant",   # Label the x axis
       y = "") + # Remove y axis label 
    scale_fill_grey()

IPEDS %>% group_by(LandGrant) %>% count(Tenure)
```

Report the summary statistic:

\vspace{0.6in}

Conditions for inference using theory-based methods for two categorical variables:

- Independence: the response for one observational unit will not influence another observational unit

\vspace{0.2in}

- Large enough sample size:

\vspace{0.7in}


Are the conditions met to analyze the university data using theory-based methods?

\vspace{0.8in}
\newpage

Steps to use theory-based methods:

* Calculate the standardized statistic

* Find the area under the standard normal distribution at least as extreme as the standardized statistic

Equation for the standard error of the difference in sample proportions assuming the null hypothesis is true:

\vspace{0.8in}

\setstretch{1.5}

* This value measures how far each possible sample difference in proportions is from the null value, on average.

\setstretch{1}

Equation for the standardized difference in sample proportions:

\vspace{0.8in}

\setstretch{1.5}

* This value measures how many standard errors the sample difference in proportions is above/below the null value.

\setstretch{1}


Calculate the standardized difference in sample proportion of higher education institutions that offer tenure between land grant universities and non-land grant universities.

* First calculate the standard error of the difference in proportion assuming the null hypothesis is true

\vspace{0.4in}


* Then calculate the Z score

\vspace{0.4in}

```{r standNormc, warning=FALSE, out.width="50%"}
openintro::normTail(m = 0, s = 1)
```

Interpret the standardized statistic

\vspace{0.5in}

\newpage

To find the p-value, find the area under the standard normal distribution at the standardized statistic and more extreme.

```{r, echo=TRUE, collapse=FALSE}
pnorm(0.985, lower.tail = FALSE)*2
```
Interpretation of the p-value:

* Statement about probability or proportion of samples

* Statistic (summary measure and value)
    
* Direction of the alternative 
    
* Null hypothesis (in context) 

\vspace{0.8in}

Conclusion with scope of inference: 

* Amount of evidence
    
* Parameter of interest 
    
* Direction of the alternative hypothesis

* Generalization

* Causation

\vspace{0.6in}
 

### Confidence interval - Video 15.3TheoryIntervals {-}

* Estimate the _______________ in true _______________ 

* $CI = \text{statistic} \pm \text{margin of error}$

#### Theory-based method for a two categorical variables {-}

* $CI = \hat{p}_1-\hat{p}_2 \pm (z^* \times SE(\hat{p}_1-\hat{p}_2))$

\setstretch{1.5}

* When creating a confidence interval, we no longer assume the _____________ hypothesis is true. Use the sample _____________ to calculate the sample to sample variability, rather than $\hat{p}_{pooled}$.

\setstretch{1}

Equation for the standard error of the difference in sample proportions *NOT* assuming the null is true:

\vspace{0.6in}

\newpage

Example:  Estimate the difference in true proportions of higher education institutions that offer tenure between land grant universities and non-land grant universities.

Find a 90\% confidence interval:

* 1st find the $z^*$ multiplier 

```{r, echo=TRUE, collapse=FALSE}
qnorm(0.95, lower.tail=TRUE)
```

* Next, calculate the standard error for the difference in proportions **NOT** assuming the null hypothesis is true

\vspace{0.8in}

* Calculate the margin of error

\vspace{0.6in}

* Calculate the endpoints of the 90\% confidence interval

\vspace{0.6in}

Confidence interval interpretation:

* How confident you are (e.g., 90%, 95%, 98%, 99%)
    
* Parameter of interest
    
* Calculated interval
    
* Order of subtraction when comparing two groups

\vspace{0.8in}

<!-- ### Relative Risk - Video RelativeRisk {-} -->

<!-- * Relative risk is the ratio of the risks in two different categories of an explanatory variable. -->

<!-- Relative Risk: -->

<!-- \vspace{0.3in} -->

<!-- Example: In a study reported in the New England Journal of Medicine [@doit2015], one-hundred fifty (150) children who had shown sensitivity to peanuts were randomized to receive a flour containing a peanut protein or a placebo flour for 2.5 years. At age 5 years, children were tested with a standard skin prick to see if they had an allergic reaction to peanut protein (yes or no). 71% of those in the peanut flour group no longer demonstrated a peanut allergy compared to 2% of those in the placebo group. -->

<!-- * Calculate the relative risk of desensitization comparing the peanut flour group to the placebo group. -->

<!-- \vspace{0.8in} -->

<!-- \setstretch{1.5} -->

<!-- * Interpretation: -->

<!--     * The proportion of successes in group 1 is the $RR$ ________________ the proportion of successes in group 2. -->

<!-- Increase in risk: -->

<!-- \vspace{0.3in} -->

<!-- * Interpretation: -->

<!--     * The proportion of successes in group 1 is the $(RR-1)$ ______________ -->
<!-- higher/lower than the proportion of successes in group 2. -->

<!-- Percent increase in risk: -->

<!-- \vspace{0.3in} -->

<!-- * Interpretation: -->

<!--     * The proportion of successes in group 1 is the $(RR-1)\times 100$ __________ higher/lower than the proportion of successes in group 2. -->

<!-- \setstretch{1} -->

<!-- * Interpret the value of relative risk from the peanut study in context of the problem. -->

<!-- \vspace{0.6in} -->

<!-- * Find the increase (or decrease) in risk of desensitization and interpret this value in context of the problem. -->

<!-- \vspace{1in} -->

<!-- * Find the percent increase (or decrease) in risk of desensitization and interpret this value in context of the problem. -->

<!-- \vspace{1in} -->

<!-- Within the peanut flour group, the percent desensitized within each age group (at start of study) is as follows: -->

<!-- 1-year-olds: 71\%; 2-year-olds: 35\%; 3-year-olds: 19\% -->

<!-- * Calculate the relative risk of desensitization comparing the 3 year olds to the 2 year olds within the peanut flour group. -->

<!-- \vspace{0.8in} -->

<!-- * Interpret the percent increase (or decrease) in risk of desensitization comparing the 3 year olds to the 2 year olds within the peanut flour group. -->

<!-- \vspace{0.8in} -->

<!-- #### Relative risk in the news {-} -->

<!-- People 50 and older who have had a mild case of covid-19 are 15\% more likely to develop shingles (herpes zoster) within six months than are those who have not been infected by the coronavirus, according to research published in the journal Open Forum Infectious Diseases [@bhavsar2022]. -->

<!-- * What was the calculated relative risk of developing shingles when comparing those who has mild COVID-19 to those who had not had COVID-19, among the 50 and older population? -->

<!-- \vspace{0.8in} -->

<!-- #### Testing Relative Risk {-} -->

<!-- In Unit 2, we tested for a difference in proportion.  We could also test for relative risk. -->

<!-- \setstretch{1.5} -->

<!-- Null Hypothesis: -->

<!-- $H_0:$ -->

<!-- \vspace{0.2in} -->

<!-- Alternative Hypothesis: -->

<!-- $H_A:$ -->

<!-- \vspace{0.2in} -->

<!-- \setstretch{1} -->

### Concept Check

Be prepared for group discussion in the next class. One member from the table should write the answers to the following on the whiteboard.

1. What conditions must be met to use the Normal Distribution to approximate the sampling distribution for the difference in sample proportions?

\vspace{0.8in}

2. Explain why a theory-based confidence interval for the Good Samaritan study from last module would NOT be similar to the bootstrap interval created.

\vspace{1in}

\newpage
