# Exploratory Data Analysis and Inference for a Quantitative Response with Independent Samples

## Vocabulary Review and Key Topics

Review the Golden Ticket posted in the resources at the end of the coursepack for a summary of a categorical explanatory variable and a quantitative response variable for independent samples.  

### Key topics

Module 11 will cover exploratory data analysis and both simulation-based and theory-based methods of inference for a quantitative response variable with independent samples. The **summary measure** for a quantitative response with independent samples is a **difference in means**. 

* Notation for a difference in sample means: $\bar{x}_1 - \bar{x}_2$, where 1 represents the 1st group of the explanatory variable and 2 represents the 2nd group

* Notation for a difference in population means: $\mu_1 - \mu_2$

Types of plots for a quantitative response variable with independent samples:

* Side-by-side boxplots

* Stacked histograms

* Stacked dotplots

R code to find the summary statistics for a quantitative response variable with independent samples:
```{r, eval=FALSE, echo=TRUE}
object %>%
  reframe(favstats(response ~ explanatory))
```


### Vocabulary

#### Plotting a quantitative response with independent groups {-}

* **Side-by-side boxplots**: plots a boxplot of the five number summary for each categorical level. R code to create side-by-side boxplots:
    ```{r, echo=TRUE, eval=FALSE}
object %>%  # Data set piped into...
  ggplot(aes(y = response, x = explanatory))+  # Identify variables
  geom_boxplot()+  # Tell it to make a box plot
  labs(title = "Don't forget to include a title",  # Title: should include the type of plot,
       # observational units, variables
       x = "x-axis label",    # x-axis label
       y = "y-axis label")  # y-axis label
    ```


\vspace{1mm}

#### Hypotheses {-}

* **Hypotheses in notation for a difference in means**: In the hypotheses below, the **null value** is equal to zero.

$$H_0: \mu_1 - \mu_2 = 0 ~~~ \text{or}~~~ H_0: \mu_1 = \mu_2 $$
$$H_A: \mu_1 - \mu_2 \left\{
\begin{array}{ll}
< \\
\ne \\
< \\
\end{array}
\right\}
0 
~~~ \text{or} ~~~ H_A:
\mu_1 \left\{
\begin{array}{ll}
< \\
\ne \\
< \\
\end{array}
\right\}
\mu_2 $$

#### Simulation-based inference for a difference in means {-}

* **Conditions necessary to use simulation-based methods for inference for a quantitative response with independent groups**:

    * **Independence**: there must be independence of observational units within groups and between groups.

* **Simulation-based methods to create the null distribution**: R code for simulation-based methods to find the p-value using the `two_mean_test` function in the `catstats` package.

    ```{r, eval=FALSE, echo=TRUE}
two_mean_test(response~explanatory, #Enter the names of the variables 
               data = object,  # Enter the name of the dataset
              first_in_subtraction = "xx", # First outcome in order of subtraction 
               number_repetitions = 10000,  # Number of simulations 
               as_extreme_as = xx,  # Observed statistic 
               direction = "xx")  # Direction of alternative: "greater", "less", or "two-sided"
    ```


* **Simulation-based methods to create the bootstrap distribution**: R code to find the simulation-based confidence interval using the `two_mean_bootstrap_CI` function from the `catstats` package.

    ```{r, eval=FALSE, echo=TRUE}
two_mean_bootstrap_CI(response ~ explanatory, #Enter the name of the variables
                      data = object,  # Enter the name of the data set
                      first_in_subtraction = "xx", # First value in order of subtraction
                      number_repetitions = 10000,  # Number of simulations
                      confidence_level = xx)
    ```

    * Review how to interpret a confidence interval for two groups from Module 8.

#### Theory-based inference for a difference in means {-}

* **Conditions for the sampling distribution of $\bar{x}_1 - \bar{x}_2$ to follow an approximate normal distribution**:

    * **Independence**: the sample’s observations are independent, e.g., are from a simple random sample and there is independence between groups. (*Remember*: This also must be true to use simulation methods!)

     * **Normality Condition**: either the sample observations come from a normally distributed population or we have a large enough sample size.  _When we have two samples, we need to check this condition for each group!_ To check this condition, use the following rules of thumb (for both $n_1$ and $n_2$):
     
         - $n < 30$: The distribution of the sample must be approximately normal with no outliers.
         
         - $30 \le n < 100$: We can relax the condition a little; the distribution of the sample must have no extreme outliers or skewness.
         
         - $n \ge 100$: Can assume the sampling distribution of $\bar{x}$ is nearly normal, even if the underlying distribution of individual observations is not.
         
* **Standard error of the sample difference in means**:
$$SE(\bar{x}_1 - \bar{x}_2) = \sqrt{\frac{{s_1}^2}{n_1}+\frac{{s_2}^2}{n_2}}$$

* **Standardized sample difference in means**:
$$T = \frac{\bar{x}_1-\bar{x}_2-0}{SE(\bar{x}_1 - \bar{x}_2)}$$
    * Use the `pt` function in R to find a theory-based p-value for a hypothesis test involving a difference in means by finding the area under a $t$-distribution with $\min(n_1-1, n_2-1)$ (the minimum sample size minus 1) degrees of freedom where $T$ is as or more extreme as the value observed (in the direction of $H_A$).

* **Margin of error**: half the width of the confidence interval. For a difference in means, the margin of error is:
$$ME = t^* \times SE(\bar{x}_1 - \bar{x}_2)$$
where $t^*$ is the **multiplier**, corresponding to the desired confidence level found from a $t$-distribution with $\min(n_1-1, n_2-1)$ degrees of freedom. 

    * Use the `qt` function in R to find the $t^*$ multiplier with $\min(n_1-1, n_2-1)$ degrees of freedom.

    * To find the endpoints of a confidence interval, add and subtract the margin of error to the sample statistic. The confidence interval for a population difference in means is:
    $$\bar{x}_1 - \bar{x}_2 \pm ME$$



\newpage