## Video Notes: Inference for Independent Samples 

Read Section 5.6 and Chapters 19 and 20 in the course textbook.  Use the following videos to complete the video notes for Module 11.

### Course Videos

* Quantitative_Response_Independent_Groups_EDA

* Simulation_Tests_Quantitative_Response_Independent_Groups

* Bootstrap_Intervals_Quantitative_Response_Independent_Groups

* Theoretical_Tests_Quantitative_Response_Independent_Groups

* Theoretical_Intervals_Quantitative_Response_Independent_Groups

\setstretch{1}

### Video: Quantitative Response Measured Across Independent Groups - Statistics and Plots (section 5.6)

\setstretch{1.5}

In this module, we will be comparing a _____________________ response variable across independent groups (a ___________________ explanatory variable).

* Independent groups:  When the measurements in one sample are not related to the measurements in the other sample.

Reminder, we use a _____________ to summarize a quantitative variable.  When comparing values between two groups, we will use ________________ to represent the two groups being compared.


Notation:

\setstretch{1.5}

* Population mean for group 1:

* Population mean for group 2:

* Parameter (difference in population means): 

* Sample mean for group 1:

* Sample mean for group 2:

* Statistic (difference in sample means): 

* Sample size for group 1:

* Sample size for group 2:

* Population standard deviation for group 1:

* Population standard deviation for group 2:

* Sample standard deviation for group 1:

* Sample standard deviation for group 2:

Plot for comparing a quantitative response variable measured across independent groups (a categorical explanatory variable):

\setstretch{1.5}

* Side-by-side boxplot: 

\rgi \rgi - One boxplot per _____________________

\rgi \rgi - Scale of the boxplot represents the ___________________________

When comparing distributions of a quantitative response variable measured across independent groups (a categorical explanatory variable), be sure to address the four characteristics discussed in Chapter 5:

* Center (reported as either the ____________ or _____________)

* Spread or variability (reported as either the ____________ or _____________)

* Shape (described as roughly symmetric, skewed-left, or skewed-right)

* Outliers (any values less than __________________________ or greater than __________________________)



### Video: Simulation Testing for a Quantitative Response Measured Across Independent Groups (section 19.1) {-}

\setstretch{1}

* The summary measure for two independent groups is the ______________________ in _________________ ________________.

#### Hypothesis Testing {-}

Conditions:

* Independence: the response for one observational unit will not influence the outcome for another observational unit

Null hypothesis assumes “no effect”, “no difference”, “nothing interesting happening”, etc.

\rgi Always of form:  “parameter” = null value

$H_0:$

\vspace{0.5in}

$H_A:$

\vspace{0.5in}

* Research question determines the alternative hypothesis.

#### Simulation-based method {-}

* Simulate many samples assuming $H_0: \mu_1 = \mu_2$

    * Write the ___________________ variable on cards

    * Mix the ___________________ variable groups together
    
    * Shuffle cards into two "new" _________________ variable groups, the sample sizes of the original groups ($n_1$ and $n_2$)

    * Calculate and plot the simulated _________________ _____ _______________ __________________ from each simulation

    * Repeat 10000 times (simulations) to create the ____________ distribution

    * Find the proportion of ___________________ at least as extreme as __________________________
    

#### Optional Notes: Additional Example {-}

Example:  Fifty-one (51) college students volunteered to look at impacts on memorization, specifically if putting letters into recognizable patterns (like FBI, CIA, EDA, CDC, etc.) would increase the number letters memorized. [@miller1956] The college students were randomly assigned to either a recognizable or non-recognizable letter group. After a period of study time, the number of letters memorized was collected on each study.  Is there evidence that putting letters into recognizable letter groups improve memory?  Use recognizable - non-recognizable as the order of subtraction.

Observational units: 

\vspace{0.2in}

Explanatory variable (include: what is group 1?): 

\vspace{0.2in}

Response variable (include: units of measure):

\vspace{0.2in}

Define the parameter in words and write it using proper notation:

\vspace{0.5in}

Write the null and alternative hypotheses in proper notation:

$H_0:$

\vspace{0.2in}

$H_A:$

\vspace{0.2in}

```{r, echo=TRUE, collapse=FALSE}
letters<-read.csv("data/letters.csv")
letters %>%
    reframe(favstats(Memorized~Grouped))
```

Calculate the summary statistic and write it using proper notation:

\vspace{0.3in}

Interpretation of the summary statistic:

\vspace{0.4in}

```{r, echo=TRUE, out.width="60%"}
letters%>%
  ggplot(aes(y = Memorized, x = Grouped))  + #Enter the name of the explanatory and response variable
  geom_boxplot()+
  labs(title = "Boxplot of Number of Letters memorized by Type 
       of Grouping for College Students", #Title your plot
       y = "Number of letters memorized", #y-axis label
       x = "Letter Grouping") #x-axis label
```


Is the independence condition met for simulation inference?

\vspace{0.4in}


Null distribution:

```{r, echo = TRUE}
set.seed(216)
two_mean_test(Memorized~Grouped, #Enter the names of the variables
              data = letters,  # Enter the name of the dataset
              first_in_subtraction = "Recognizable", # First outcome in order of subtraction
              number_repetitions = 10000,  # Number of simulations
              as_extreme_as = 3.166,  # Observed statistic
              direction = "greater")  # Direction of alternative: "greater", "less", or "two-sided"
```

Explain why the null distribution is centered at the value of zero:

\vspace{1in}

Where is the red line plotted on the null distribution?  Why is that value important?

\vspace{0.3in}

Why is the area right of the red line shaded on the null distribution?

\vspace{0.3in}

What is the p-value of the test?

\vspace{0.2in}

Interpretation of the p-value:

* Statement about probability or proportion of samples

* Statistic (summary measure and value) and Direction of the alternative 
    
* Null hypothesis (population reference, summary measure, equal to null value)

* Context of the problem (observational units, variables (for a quantitative response measured across independent groups, include: both explanatory variable groups, units of measure of the response variable, and order of subtraction))

\vspace{0.8in}

Conclusion with scope of inference: 

* Amount of evidence
    
* For the alternative hypothesis (population reference, summary measure, direction, null value)

* Context of the problem (observational units, variables (for a quantitative response measured across independent groups, include: both explanatory variable groups, units of measure of the response variable, and order of subtraction))

* Generalization

* Causation

\vspace{0.8in}


### Video: Bootstrap Confidence Interval for a Quantitative Response Measured Across Independent Groups (section 19.2) {-}

To estimate the difference in true mean, we will create a confidence interval.


#### Simulation-based method {-}

* Write the response variable value on cards

* Keep explanatory variable groups ______________
    
* Sample with replacement __________ times in explanatory variable group 1 and ________ times in explanatory variable group 2

* Calculate and plot the simulated _________________ _____ _______________ __________________ from each simulation

* Repeat 10000 times (simulations) to create the _________________ distribution

* Find the cut-offs for the middle X\% (confidence level) in a bootstrap distribution.


#### Optional Notes: Additional Example {-}

For the letters example, we will estimate the difference in true mean number of letters recognized for students given recognizable letter groupings and students given non-recognizable letter groupings.


Bootstrap distribution:

```{r, echo = TRUE}
set.seed(216)
two_mean_bootstrap_CI(Memorized ~ Grouped, #Enter the name of the variables
                      data = letters,  # Enter the name of the data set
                      first_in_subtraction = "Recognizable", # First value in order of subtraction
                      number_repetitions = 10000,  # Number of simulations
                      confidence_level = 0.95)
```

Why is the bootstrap distribution centered near 3.2?  Why is that number important?

\vspace{0.4in}

Confidence interval interpretation:

* How confident you are (e.g., 90%, 95%, 98%, 99%)
    
* Parameter of interest (including context: observational units, variables (for a quantitative response measured across independent groups, include: both explanatory variable groups, units of measure of the response variable, and order of subtraction))
    
* Calculated interval

\vspace{0.8in}


### Video: Theoretical Testing for a Quantitative Response Measured Across Independent Groups (section 19.3.1) {-}

Conditions:

* Independence: the response for one observational unit will not influence the outcome for another observational unit (in either group)

* Large enough sample size

\vspace{1in}

Like with a single quantitative variable, the $t$-distribution can be used to model the difference in means. 

\setstretch{1.5}

* For independent groups, we use the ______- distribution with ________________ degrees of freedom to approximate the sampling distribution.

\setstretch{1}

General steps of a hypothesis test

1.	Write a research question and hypotheses.

2.	Collect data and calculate a summary statistic.

3.	Model a sampling distribution which assumes the null hypothesis is true.

\rgi \rgi - If the conditions are met, the sampling distribution of a difference in sample means is the $t$-distribution with smaller sample size $-1$ degrees of freedom.

4.	Calculate a p-value.

\rgi \rgi - Calculate the standardized difference in sample means ($T$) and compare that to the $t$-distribution with smaller sample size $-1$ degrees of freedom.
\rgi \rgi - P-value will be the area under the curve beyond $T$ (in the direction of the alternative hypothesis)

5.	Draw conclusions based on a p-value.


Equation for the standard error of the difference in sample means:

\vspace{0.5in}

Equation for the standardized difference in sample means (standardized statistic):

\vspace{0.5in}

 
#### Optional Notes: Additonal Example {-}

Every year, orange and black monarch butterflies migrate from their summer breeding grounds in the US and Canada to mountain forests in central Mexico, where they hibernate for the winter. Due to abnormal weather patterns and drought affecting monarch habitats and feeding grounds, the population of monarch butterflies is estimated to have decreased by 53% from the 2018-2019 wintering season to the 2019-2020 wintering season (WWF, 2020). While conservationists often resort to captive-rearing with the goal of raising biologically indistinct individuals for release into the wild, tagging studies have shown that captive-reared monarchs have lower migratory success compared to wild monarchs. For this study, the researchers raised 67 monarchs (descended from wild monarchs) from eggs to maturity and then compared them to a group of 40 wild-caught monarchs. The researchers want to explore whether the maximum grip strength (how many Newtons a butterfly exerts at the moment of release when gently tugged from a mesh-covered perch) differs between captive-reared and wild-caught monarchs. Use Captive – Wild for order of subtraction.

Observational units: 

\vspace{0.2in}

Explanatory variable (include: what is group 1?): 

\vspace{0.2in}

Response variable (include: units of measure):

\vspace{0.2in}

Define the parameter in words and write it using proper notation:

\vspace{0.5in}

Write the null and alternative hypotheses in notation.

$H_0:$

\vspace{0.2in}

$H_A:$

\vspace{0.2in}


```{r, echo=TRUE, eval=TRUE, collapse=FALSE}
butterfly <-read.csv("data/butterfly1.csv")
```

```{r, echo=FALSE}
butterflies <- butterfly %>% na.omit() %>%
    rename(Monarch_Group = "Monarch.Group",
           MaxGrip = "Max.Grip.Strength..N.") %>%
    mutate(Monarch_Group = factor(Monarch_Group),
           Sex = factor(Sex)) %>%
    mutate(Monarch_Group = fct_collapse(Monarch_Group,"Captive" = c("Incubator - Fall conditions", "Rearing room - summer conditions"), "Wild" = "Wild migrants"))
```

```{r, echo=TRUE}
butterflies %>%
    reframe(favstats(MaxGrip~Monarch_Group))
```

```{r}
butterflies%>%
    ggplot(aes(y = MaxGrip, x = Monarch_Group)) + #Enter the name of the explanatory and response variable
    geom_boxplot()+
    labs(title = "Boxplot of Max Grip of Monarch Butterflies by 
         Environment Reared", #Title your plot
         y = "Max Grip (Newtons)", #y-axis label
         x = "Environment Reared") #x-axis label
```


Are the conditions met to analyze the butterfly data using theory based-methods?

\vspace{0.8in}

Calculate the standard error of the difference in sample means: $SE(\bar{y}_1 - \bar{y}_2)$

\vspace{0.6in}

Calculate the standardized difference in sample means (standardized statistic):

\vspace{1in}

Interpret the standardized statistic.

\vspace{0.5in}

Label the standardized difference in sample means (standardized statistic) on the theoretical distribution below and shade the area representing the p-value.

```{r, tstarpb2, echo = F}

x <- seq(-5, 5, length=100)
hx<-dt(x, 39)
degf <- 39

plot(x, hx, type="l", lty=1, lwd=3, xlab="",
  ylab="Density", main="t-distribution with 39 df")
```
Why did the theoretical $t$-distribution used have 39 degrees of freedom?

\vspace{0.3in}

To find the theory-based p-value:

```{r, echo=TRUE, collapse=FALSE}
pt(-5, df=39, lower.tail=TRUE)*2
```

Interpretation of the p-value:

* Statement about probability or proportion of samples

* Statistic (summary measure and value) and Direction of the alternative 
    
* Null hypothesis (population reference, summary measure, equal to null value)

* Context of the problem (observational units, variables (for a quantitative response measured across independent groups, include: both explanatory variable groups, units of measure of the response variable, and order of subtraction))


\vspace{0.8in}

Conclusion with scope of inference: 

* Amount of evidence
    
* For the alternative hypothesis (population reference, summary measure, direction, null value)

* Context of the problem (observational units, variables (for a quantitative response measured across independent groups, include: both explanatory variable groups, units of measure of the response variable, and order of subtraction))

* Generalization

* Causation

\vspace{0.8in}



###  Theoretical Confidence Intervals for a Quantitative Response Measured Across Independent Groups (section 19.3.2){-}

General formula for a confidence interval is always:

\rgi \rgi $\text{statistic} \pm \text{margin of error}$

\rgi \rgi where $\text{margin of error} = \text{multiplier} \times \text{standard error of the statistic}$

For a quantitative response variable measured across independent groups,

* $\text{statistic}$ = ___________

* $\text{margin of error} = \text{multiplier} \times \text{standard error of the statistic} = $ ____________ $\times$ ______________

The formula for the standard error of the difference in sample means is the same as was used in the previous video: $SE(\overline{y_1} - \overline{y_2}) = \sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}$

The $t^*$ multiplier is the value at the given percentile(s) of the t-distribution with smaller sample size $- 1$ degrees of freedom.

#### Optional Notes: Additional Example {-}

Using the butterfly data, calculate the 99\% confidence interval.

```{r, echo=TRUE, collapse=FALSE}
butterflies %>%
    reframe(favstats(MaxGrip~Monarch_Group))
```

* Need the $t^*$ multiplier for a 90% confidence interval from a t-distribution with 39 df.

```{r, echo=TRUE, collapse=FALSE}
qt(0.95, df=39, lower.tail = TRUE)
```

Draw a line at the provided $t^\star$ on the $t$-distribution with 39 degrees of freedom shown below, and label the percentile of the $t^\star$ value.  

```{r, tstarpb2, echo = F}

x <- seq(-5, 5, length=100)
hx<-dt(x, 39)
degf <- 39

plot(x, hx, type="l", lty=1, lwd=3, xlab="",
  ylab="Density", main="t-distribution with 39 df")

```

* We will use the same value for the $SE(\bar{y}_1-\bar{y}_2)$ as calculated for the standardized statistic.

\vspace{1in}

Calculate the margin of error for a 90\% confidence interval for the parameter of interest.

\vspace{0.5in}

Calculate a 90\% confidence interval for the parameter of interest.

\vspace{0.6in}

* How confident you are (e.g., 90%, 95%, 98%, 99%)
    
* Parameter of interest (including context: observational units, variables (for a quantitative response measured across independent groups, include: both explanatory variable groups, units of measure of the response variable, and order of subtraction))
    
* Calculated interval

\vspace{0.8in}




### Concept Check

Be prepared for group discussion in the next class. One member from the table should write the answers to the following on the whiteboard.

1. Why is the recognizable letter study analyzed as two independent groups rather than paired data?

\vspace{0.6in}

2. Write out the equation for the standard error for a difference in sample means.

\vspace{1in}

\newpage
