## Lecture Notes Week 12: Inference for Independent Samples 

\setstretch{1}

### Single categorical, single quantitative variable with independent samples {-}

\setstretch{1.5}

* In this week, we will study inference for a ______________________ explanatory variable and a _________________________ response variable where the two groups are ____________________________.

* Independent groups:  When the measurements in one sample are not
___________________ to the measurements in the other sample.

\setstretch{1}

* Two random samples taken separately from two populations and the same response variable is recorded. Compare the average number of sick days off from work for people who had a flu shot and people who didn’t.

* Participants are randomly assigned to one of two treatment conditions, and the same response variable is recorded.

Rather than analyzing the differences as a single mean we will calculate summary statistics on each sample.  

\setstretch{1.5}

* The summary measure for two independent groups is the ______________________ in _____________________________.

* Difference in means: the difference in average _______________________ variable outcome for observational units between ____________________ variable groups

\setstretch{1}

Parameter of Interest:  

* Include:
 
    * Reference of the population (true, long-run, population, all)
	
	* Summary measure

	* Context
	   
	    * Observational units/cases
	   
	    * Response variable (and explanatory variable if present)
	       
	        * If the response variable is categorical, define a ‘success’ in context

$\mu_1-\mu_2:$ 

\vspace{0.5in}

\setstretch{1.5}

Notation for the Sample Statistics

* Sample mean for group 1:

* Sample mean for group 2:

* Sample difference in means: 

* Sample standard deviation for group 1:

* Sample standard deviation for group 2:

* Sample size for group 1:

* Sample size for group 2:

\setstretch{1}

Example:  Fifty-one (51) college students volunteered to look at impacts on memorization, specifically if putting letters into recognizable patterns (like FBI, CIA, EDA, CDC, etc.) would increase the number letters memorized.  The college students were randomly assigned to either a recognizable or non-recognizable letter group. After a period of study time, the number of letters memorized was collect on each study.  Is there evidence that putting letters into recognizable letter groups improve memory?

Why should we treat this as two independent groups rather than paired data?

\vspace{0.6in}

Parameter of interest:

\vspace{0.8in}

```{r, echo=TRUE}
letters<-read.csv("data/letters.csv")
letters %>%
    reframe(favstats(Memorized~Grouped))
```

Summary statistic:

\vspace{0.4in}

```{r, echo=TRUE}
letters%>%
  ggplot(aes(y = Memorized, x = Grouped))  + #Enter the name of the explanatory and response variable
  geom_boxplot()+
  labs(title = "Boxplot of number of letters memorized by type of grouping", #Title your plot
       y = "Number of letter memorized", #y-axis label
       x = "Letter Grouping") #x-axis label
```

### Hypothesis Testing {-}

Conditions:

* Independence:

\vspace{0.3in}

Null hypothesis assumes “no effect”, “no difference”, “nothing interesting happening”, etc.

\rgi Always of form:  “parameter” = null value

$H_0:$

\vspace{0.5in}

$H_A:$

\vspace{0.5in}

* Research question determines the alternative hypothesis.

Write the null and alternative hypotheses for the letters study:

In words: 

$H_0:$

\vspace{0.5in}

$H_A:$

\vspace{0.5in}

In notation:

$H_0:$

\vspace{0.2in}

$H_A:$

\vspace{0.2in}

#### Simulation-based method {-}

* Simulate many samples assuming $H_0: \mu_1 = \mu_2$

    * Write the response variable values on cards

    * Mix the explanatory variable groups together
    
    * Shuffle cards into two explanatory variable groups to represent the sample size in each group ($n_1$ and $n_2$)

    * Calculate and plot the simulated difference in sample means from each simulation

    * Repeat 1000 times (simulations) to create the null distribution

    * Find the proportion of simulations at least as extreme as $\bar{x}_1 - \bar{x}_2$
    
```{r, echo = TRUE}
set.seed(216)
two_mean_test(Memorized~Grouped, #Enter the names of the variables
              data = letters,  # Enter the name of the dataset
              first_in_subtraction = "Recognizable", # First outcome in order of subtraction
              number_repetitions = 1000,  # Number of simulations
              as_extreme_as = 3.166,  # Observed statistic
              direction = "greater")  # Direction of alternative: "greater", "less", or "two-sided"
```

Explain why the null distribution is centered at the value of zero:

\newpage

Interpretation of the p-value:

* Statement about probability or proportion of samples

* Statistic (summary measure and value)
    
* Direction of the alternative 
    
* Null hypothesis (in context) 

\vspace{0.8in}

Conclusion: 

* Amount of evidence
    
* Parameter of interest 
    
* Direction of the alternative hypothesis

\vspace{0.6in}

#### Theory-based method {-}

Conditions:

* Independence

* Large enough sample size

\vspace{1in}

Like with paired data the t-distribution can be used to model the difference in means. 

\setstretch{1.5}

* For a independent samples we use the ______- distribution
with ________________ degrees of freedom to approximate the sampling distribution.

\setstretch{1}

Theory-based test:

* Calculate the standardized statistic

* Find the area under the t-distribution with the smallest $n - 1$ df [min($n_1-1, n_2-1$) at least as extreme as the standardized statistic

Equation for the standard error of the difference in sample mean:

\vspace{0.5in}

Equation for the standardized difference in sample mean:

\vspace{0.5in}

Example:

Every year, orange and black monarch butterflies migrate from their summer breeding grounds in the US and Canada to mountain forests in central Mexico, where they hibernate for the winter. Due to abnormal weather patterns and drought affecting monarch habitats and feeding grounds, the population of monarch butterflies is estimated to have decreased by 53\% from the 2018-2019 wintering season to the 2019-2020 wintering season (WWF, 2020). While conservationists often resort to captive-rearing with the goal of raising biologically indistinct individuals for release into the wild, tagging studies have shown that captive-reared monarchs have lower migratory success compared to wild monarchs. For this study, the researchers raised 67 monarchs (descended from wild monarchs) from eggs to maturity and then compared them to a group of 40 wild-caught monarchs. The researchers want to explore whether the maximum grip strength (how many Newtons a butterfly exerts at the moment of release when gently tugged from a mesh-covered perch) differs between captive-reared and wild-caught monarchs. Use Captive – Wild for order of subtraction.

Write the null and alternative hypotheses in notation.

$H_0:$ 

\vspace{0.2in}

$H_A:$

\vspace{0.2in}

```{r, echo=TRUE}
butterfly <-read.csv("data/butterfly1.csv")
butterflies <- butterfly %>% 
  na.omit() %>%
  rename(Monarch_Group = "Monarch.Group",
         MaxGrip = "Max.Grip.Strength..N.") %>%
  mutate(Monarch_Group = factor(Monarch_Group),
         Sex = factor(Sex)) %>%
  mutate(Monarch_Group = fct_collapse(Monarch_Group,
               "Captive" = c("Incubator - Fall conditions",
                             "Rearing room - summer conditions"),
               "Wild" = "Wild migrants"))

butterflies %>%
    reframe(favstats(MaxGrip~Monarch_Group))
```

```{r, echo=TRUE}
butterflies%>%
  ggplot(aes(y = MaxGrip, x = Monarch_Group))  + 
    #Enter the name of the explanatory and response variable
  geom_boxplot()+
  labs(title = "Boxplot of max grip of Monarch butterflies by environment reared", 
       #Title your plot
       y = "Max Grip (Newtons)", #y-axis label
       x = "Environment Reared") #x-axis label
```
Are the conditions met to analyze the data using theory based-methods?

\vspace{1in}

Calculate the standardized difference in max grip strength.

* First calculate the $SE(\bar{x}_1 - \bar{x}_2)$

\vspace{0.6in}

* Then calculate the T-score

\vspace{1in}

```{r, pvaluepb, echo = F}

x <- seq(-4, 4, length=100)
hx<-dt(x, 39)
degf <- 39

plot(x, hx, type="l", lty=1, lwd=3, xlab="",
  ylab="Density", main="t-distribution with 39 df")

```

Interpret the standardized statistic:

\vspace{0.8in}

To find the theory-based p-value:

```{r, echo=TRUE}
pt(-5, df=39, lower.tail=TRUE)*2
```

### Confidence interval {-}

To estimate the difference in true mean we will create a confidence interval.

#### Simulation-based method {-}

* Write the response variable values on cards

* Keep explanatory variable groups separate
    
* Sample with replacement $n_1$ times in explanatory variable group 1 and $n_2$ times in explanatory variable group 2

* Calculate and plot the simulated difference in sample means from each simulation

* Repeat 1000 times (simulations) to create the bootstrap distribution

* Find the cut-offs for the middle X\% (confidence level) in a bootstrap distribution.

Returning to the letters example, we will estimate the difference in true mean number of letters recognized for students given  recognizable letter groupings and students given non-recognizable letter groupings.

```{r, echo = TRUE}
set.seed(216)
two_mean_bootstrap_CI(Memorized ~ Grouped, #Enter the name of the variables
                      data = letters,  # Enter the name of the data set
                      first_in_subtraction = "Recognizable", # First value in order of subtraction
                      number_repetitions = 1000,  # Number of simulations
                      confidence_level = 0.95)
```

Confidence interval interpretation:

* How confident you are (e.g., 90\%, 95\%, 98\%, 99\%)
    
* Parameter of interest
    
* Calculated interval
    
* Order of subtraction when comparing two groups

\vspace{0.8in}


#### Theory-based method {-}

* Calculate the interval centered at the sample statistic

\rgi $\text{statistic} \pm \text{margin of error}$

\vspace{0.8in}

\newpage

Using the Monarch butterfly data, calculate the 99\% confidence interval.

```{r, echo=TRUE}
butterflies %>%
    reframe(favstats(MaxGrip~Monarch_Group))
```

* Need the $t^*$ multiplier for a 99\% confidence interval from a t-distribution with _________ df.

```{r, echo=TRUE}
qt(0.995, df=39, lower.tail = TRUE)
```

* We will use the same value for the $SE(\bar{x}_1-\bar{x}_2)$ as calculated for the standardized statistic.

\vspace{1in}

Based on the p-value and confidence interval, write a conclusion to the test.

\vspace{1in}
\newpage
