## Video Notes: Inference for a single quantitative variable 

Read Chapters 5 and 17 in the course textbook.  Use the following videos to complete the video notes for Module 6.

### Course Videos

* QuantitativeData

<!-- * 18.1and18.2 -->

<!-- * 18.3 -->

\setstretch{1}


### Single quantitative variable

* Reminder: review summary measures and plots discussed in the Week 3 material and Chapter 5 of the textbook.

* The summary measure for a single quantitative variable is the ______________.

\setstretch{1.5}

Notation:

- Population mean:

- Population standard deviation:

- Sample mean:

- Sample standard deviation:

- Sample size:

\setstretch{1}

Example:  What is the average weight of adult male polar bears?  The weight was measured on a representative sample of 83 male polar bears from the Southern Beaufort Sea.

```{r, echo=TRUE}
pb <- read.csv("https://math.montana.edu/courses/s216/data/polarbear.csv")
```

Plots of the data:

```{r, echo=TRUE, out.width="60%"}
pb %>%
    ggplot(aes(x = Weight)) +   # Name variable to plot
    geom_histogram(binwidth = 10) +  # Create histogram with specified binwidth
    labs(title = "Histogram of Male Polar Bear Weight", # Title for plot
       x = "Weight (kg)", # Label for x axis
       y = "Frequency") # Label for y axis

pb %>% # Data set piped into...
ggplot(aes(x = Weight)) +   # Name variable to plot
  geom_boxplot() +  # Create boxplot
  labs(title = "Boxplot of Male Polar Bear Weight", # Title for plot
       x = "Weight (kg)", # Label for x axis
       y = "Frequency") # Label for y axis
```

Summary Statistics:

```{r, echo=TRUE}
pb %>%
  summarise(favstats(Weight)) #Gives the summary statistics
```
### Confidence interval {-}

#### Simulation-based method{-}

* Label cards with the values from the data set

* Sample with replacement (bootstrap) from the original sample $n$ times

* Plot the simulated sample mean on the bootstrap distribution

* Repeat at least 1000 times (simulations)

* Find the cut-offs for the middle X% (confidence level) in a bootstrap distribution.

* ie. 95% CI = (2.5th percentile, 97.5th percentile)

Conditions for inference for a single mean:

- Independence:

\vspace{0.5in}

```{r, echo=TRUE, warning=FALSE}
set.seed(216)
one_mean_CI(pb$Weight,
  summary_measure = "mean",
  number_repetitions = 10000,
  confidence_level = 0.95)
```
The confidence interval estimates the ________________
of ____________________.

Confidence interval interpretation:

* How confident you are (e.g., 90%, 95%, 98%, 99%)

* Parameter of interest

* Calculated interval

* Order of subtraction when comparing two groups

\vspace{0.8in}

\newpage

#### Theory-based method {-}

* Calculate the interval centered at the sample statistic

\rgi $\text{statistic} \pm \text{margin of error}$

\vspace{0.5in}

Conditions for inference using theory-based methods:

- Independence:

\vspace{0.2in}

- Large enough sample size:

\vspace{0.2in}

### T - distribution {-}

In the theoretical approach, we use the CLT to tell us that the distribution of sample means will be approximately normal, centered at the assumed true mean under $H_0$ and with standard deviation $\frac{\sigma}{\sqrt{n}}$.

$$\bar{x} \sim N(\mu_0, \frac{\sigma}{\sqrt{n}})$$
\setstretch{1.5}

* Estimate the population standard deviation, $\sigma$, with the
___________________________ standard deviation, ________.

* For a single quantitative variable we use the ____ - distribution
with _______________
degrees of freedom to approximate the sampling distribution.

\setstretch{1}

The $t^*$ multiplier is the value at the given percentile of the t-distribution with $n - 1$ degrees of freedom.

```{r, tstarpb, echo = F}

x <- seq(-4, 4, length=100)
hx<-dt(x, 82)
degf <- 82

plot(x, hx, type="l", lty=1, lwd=3, xlab="",
  ylab="Density", main="t-distribution with 82 df")

```

\newpage
To find the $t^*$ multiplier for a 95\% confidence interval:

```{r, echo=TRUE}
qt(0.975, df = 82)
```
Calculation of the confidence interval for the true mean weight of polar bears from the Southern Beaufort Sea:

\vspace{0.8in}

### Hypothesis testing {-}

\setstretch{1.5}

* Hypotheses are always written about the _________________________.  For a single mean we will use the notation ___________.

\setstretch{1}

Null Hypothesis:

$H_0:$

\vspace{0.2in}
Alternative Hypothesis:

$H_A:$

\vspace{0.2in}

\setstretch{1.5}

* Direction of the alternative depends on the __________________
___________________.

\setstretch{1}

#### Simulation-based method {-}

* Simulate many samples assuming $H_0: \mu = \mu_0$

    * Shift the data by the difference between $\mu_0$ and $\bar{x}$

    * Sample with replacement $n$ times from the shifted data

    * Plot the simulated shifted sample mean from each simulation

    * Repeat 1000 times (simulations) to create the null distribution

    * Find the proportion of simulations at least as extreme as $\bar{x}$

Example:  Is there evidence that male polar bears weigh less than 370kg (previously recorded measure), on average?  The weight was measured on a representative sample of 83 male polar bears from the Southern Beaufort Sea.

Hypotheses:

In notation:

$H_0:$

\vspace{0.2in}

$H_A:$

\vspace{0.2in}

\newpage

In words:

$H_0:$

\vspace{0.6in}

$H_A:$

\vspace{0.6in}

Reminder of summary statistics:

```{r, echo=TRUE}
pb %>%
  summarise(favstats(Weight)) #Gives the summary statistics
```
Find the difference:

$\mu_0 - \bar{x} =$

```{r}
one_mean_test(pb$Weight,
  summary_measure = "mean",
  shift = 45.4,
  as_extreme_as = 324.6,
  direction = "less",
  number_repetitions = 10000,
  add_normal = FALSE)
```

\newpage

Interpretation of the p-value:

* Statement about probability or proportion of samples

* Statistic (summary measure and value)

* Direction of the alternative

* Null hypothesis (in context)


\vspace{0.8in}

Conclusion:

* Amount of evidence

* Parameter of interest

* Direction of the alternative hypothesis

\vspace{0.8in}

#### Theory-based method {-}

* Calculate the standardized statistic

* Find the area under the t-distribution with $n - 1$ df at least as extreme as the standardized statistic

Equation for the standard error of the sample mean:

\vspace{0.5in}

Equation for the standardized sample mean:

\vspace{0.5in}

Calculate the standardized sample mean weight of adult male polar bears:

\vspace{0.4in}

```{r, pvaluepb, echo = F}

x <- seq(-4, 4, length=100)
hx<-dt(x, 82)
degf <- 82

plot(x, hx, type="l", lty=1, lwd=3, xlab="",
  ylab="Density", main="t-distribution with 82 df")

```

Interpret the standardized sample mean weight:

\vspace{0.8in}

To find the theory-based p-value:

```{r, echo=TRUE}
pt(-4.683, df=82, lower.tail=TRUE)
```

\newpage
