## Activity 8:  The Good Samaritan (continued) --- Simulation-based Hypothesis Test & Confidence Interval

\setstretch{1}

### Learning outcomes

* Identify the parameter of interest for a difference in proportions.

* Create and interpret a simulation-based confidence interval for a difference in proportions.

### Terminology review

In today's activity, we will use simulation methods to estimate the difference in two proportions. Some terms covered in this activity are:

* Parameter of interest

* Bootstrapping

* Confidence interval

* Types of errors

To review these concepts, see Chapter 15 in your textbook.

### The Good Samaritan

In the out of class activity, we began a test of significance to see if people in a hurry are less likely to help those in need. Today we will use RStudio to continue to assess this research question.

Researchers at the Princeton University wanted to investigate influences on behavior [@darley1973].  The researchers randomly selected 67 students from the Princeton Theological Seminary to participate in a study.  Only 47 students chose to participate in the study, and the data below includes 40 of those students (7 students were removed from the study for various reasons).  As all participants were theology majors planning a career as a preacher, the expectation was that all would have a similar disposition when it comes to helping behavior.  Each student was then shown a 5-minute presentation on the Good Samaritan, a parable in the Bible which emphasizes the importance of helping others.  After the presentation, the students were told they needed to give a talk on the Good Samaritan parable at a building across campus.  Half the students were told they were late for the presentation; the other half told they could take their time getting across campus (the condition was randomly assigned).  On the way between buildings, an actor pretending to be a homeless person in distress asked the student for help.  The researchers recorded whether the student helped the actor or not.  The results of the study are shown in the table below.  Do these data provide evidence that those in a hurry will be less likely to help people in need in this situation?  Use the order of subtraction hurry – no hurry.

|                    | Hurry Condition | No Hurry Condition | Total |
|--------------------|-----------------|--------------------|-------|
| Helped Actor       |        2        |         11         |   13  |
| Did Not Help Actor |        18       |          9         |   27  |
| Total              |        20       |         20         |   40  |

1.  Simulate one sample using the cards provided by your instructor.  Write down the value of the simulated statistic.  How does the value of your group's simulated statistic compare to the other groups at your table?  Are the simulated values closer to the null value of zero than the actual calculated difference in proportions?

\vspace{1in}

To create the null distribution of differences in sample proportions, we will use the `two_proportion_test()` function in R (in the `catstats` package).  We will need to enter the response variable name and the explanatory variable name for the formula, the data set name (identified above as `good`), the outcome for the explanatory variable that is first in subtraction, number of repetitions, the outcome for the response variable that is a success (what the numerator counts when calculating a sample proportion), and the direction of the alternative hypothesis.

The response variable name is `Behavior` and the explanatory variable name is `Condition`.

\newpage

2.  What inputs should be entered for each of the following to create the simulation?
\vspace{1mm}

* First in subtraction (What is the outcome for the explanatory variable that is used as first in the order of subtraction? `"Hurry"` or `"No hurry"`):

\vspace{.15in}
* Number of repetitions:

\vspace{.15in}
* Response value numerator (What is the outcome for the response variable that is considered a success? `"Help"` or `"No help"`):

\vspace{.15in}
* As extreme as (enter the value for the sample difference in proportions):

\vspace{.15in}
* Direction (`"greater"`, `"less"`, or `"two-sided"`):

\vspace{.15in}

Using the R script file for this activity, enter your answers for question 16 in place of the `xx`'s to produce the null distribution with 1000 simulations; highlight and run lines 1--16.

```{r, echo=TRUE, eval=FALSE}
two_proportion_test(formula = Behavior~Condition, # response ~ explanatory
    data = good, # Name of data set
    first_in_subtraction = "xx", # Order of subtraction: enter the name of Group 1
    number_repetitions = 1000, # Always use a minimum of 1000 repetitions
    response_value_numerator = "xx", # Define which outcome is a success
    as_extreme_as = xx, # Calculated observed statistic (difference in sample proportions)
    direction="xx") # Alternative hypothesis direction ("greater","less","two-sided")
```

3.  Sketch the null distribution created here.

\vspace{1.5in}


4. What value is the null distribution centered around?  Explain why this makes sense.

\vspace{.8in}

5.  What is the value of the p-value? *Remember*: This is the value given at the bottom of the null distribution.

\vspace{0.2in}
\newpage

6.  Interpret the p-value in context of the study.

\vspace{1in}

7.  How much evidence does the p-value provide against the null hypothesis? *Hint*: Refer to the guidelines given in Activity 6A.

\vspace{0.4in}

8.  Write a conclusion to the test.

\vspace{1in}

9.  Do you expect the null value to be in a 99\% confidence interval?  Explain your answer.

\vspace{0.8in}

#### Use statistical analysis methods to draw inferences from the data {-}

In this part of the activity, we will estimate the difference in true proportion of people who will help others for those in the hurry condition and those not in the hurry condition by finding a confidence interval.

10.  Write the parameter of interest in context of the study. Use proper notation. 

\vspace{1in}

We will use the `two_proportion_bootstrap_CI()` function in R (in the `catstats` package) to simulate the bootstrap distribution of differences in sample proportions and calculate a confidence interval. We will need to enter the response variable name and the explanatory variable name for the formula, the data set name (identified above as `good`), the outcome for the explanatory variable that is first in subtraction, number of repetitions, the outcome for the response variable that is a success (what the numerator counts when calculating a sample proportion), and the confidence level as a decimal.

The response variable name is `Behavior` and the explanatory variable name is `Condition`.

11. What values should be entered for each of the following into the simulation to create a 99\% confidence interval?
\vspace{.5mm}

* First in subtraction (What is the outcome for the explanatory variable that is used as first in the order of subtraction? `"Hurry"` or `"No hurry"`):

\vspace{.15in}
* Response value numerator (What is the outcome for the response variable that is considered a success? `"Help"` or `"No help"`):

\vspace{.15in}
* Number of repetitions:

\vspace{.15in}
* Confidence level (entered as a decimal):

\vspace{.15in}

Using the R script file for this activity, enter your answers for question 7 in place of the `xx`'s to produce the bootstrap distribution with 1000 simulations; highlight and run lines 16--21.

```{r, echo=TRUE, eval=FALSE}
two_proportion_bootstrap_CI(formula = Behavior ~ Condition, 
        data=good, # Name of data set
        first_in_subtraction = "xx", # Order of subtraction: enter the name of Group 1
        response_value_numerator = "xx", # Define which outcome is a success 
        number_repetitions = 1000, # Always use a minimum of 1000 repetitions
        confidence_level = xx) # Enter the level of confidence as a decimal
```
\newpage 

12.  Where is the bootstrap distribution centered?  Explain why.

\vspace{0.8in}

13. Report the bootstrap 99\% confidence interval. 

\vspace{0.4in}

14. What percentile of the bootstrap distribution does the upper value of the confidence interval represent?

\vspace{0.3in}

15. Interpret the 99\% confidence interval in context of the problem.

\vspace{1in}

16.  What conclusion to the research question can be made based on the 99\% confidence interval?

\vspace{0.8in}

17.  Is the null value in the 99\% confidence interval? 

\vspace{0.8in}

\newpage

<!-- #### Types of errors {-} -->

<!-- Recall from a previous activity, hypothesis tests are not flawless. In a hypothesis test, there are two competing hypotheses: the null and alternative. We make a decision about which might be true, but we may choose incorrectly.   -->

<!-- <!-- |       |            | Test Conclusion |                     | --> -->
<!-- <!-- |       |            | Reject $H_0$    | Fail to reject $H_0$| --> -->
<!-- <!-- |:-----:|:----------:|-----------------|---------------------| --> -->
<!-- <!-- | Truth | $H_0$ true |  good decision  |  Type 1 Error       | --> -->
<!-- <!-- |       | $H_A$ true |  Type 2 Error   |  good decision      | --> -->

<!-- \begin{table} -->
<!-- \caption{Four different possible scenarios for hypothesis test decisions.} -->
<!-- \centering -->
<!-- \begin{tabular}[h]{ll|cc} -->
<!-- \hline -->
<!--  & &  \multicolumn{2}{c}{\textbf{Test conclusion}} \\ -->
<!--  &  & \multicolumn{1}{c}{Fail to reject $H_0$} & \multicolumn{1}{c}{Reject $H_0$}\\ -->
<!-- \hline -->
<!--  & $H_0$ true & Good decision & Type 1 Error\\ -->
<!-- \hline -->
<!-- \textbf{Truth} & $H_A$ true & Type 2 Error & Good decision\\ -->
<!-- \hline -->
<!-- \end{tabular} -->
<!-- \label{tab:errors} -->
<!-- \end{table} -->

<!-- Shown in Table \@ref(tab:errors), a **Type 1 Error** happens when we reject the null hypothesis when $H_0$ is actually true. A **Type 2 Error** happens when we fail to reject the null hypothesis when the alternative is actually true. -->

<!-- 14.  Using a significance level of 0.01 and your answer to question 13, what decision do you make in regards to the null hypothesis? -->

<!-- \vspace{0.3in} -->

<!-- 15. What type of error could we have made? -->

<!-- \vspace{0.3in} -->

<!-- 16.  Write this error in context of the problem. -->

<!-- \vspace{0.5in} -->

### Take-home messages

1.  We use the same guidelines for the strength of evidence as we did in Activity 6A.

2. To create one simulated sample on the bootstrap distribution for a difference in sample proportions, label $n_1 + n_2$ cards with the outcomes for the original responses.  Keep groups separate and randomly draw with replacement $n_1$  times from group 1 and $n_2$ times from group 2.  Calculate and plot the resampled difference in the proportion of successes. 

3. If the null value is not contained in a 99\% confidence interval, then there is evidence against the null hypothesis and the p-value is less than the significance level of 0.01.


### Additional notes

Use this space to summarize your thoughts and take additional notes on today's activity and material covered.

\newpage
